# The idea {#sec-idea}

```{r}
#| label: setup
source("R/graphemb.R")
set.seed(1)
```

## The problem

Here is a graph. Zachary's karate club [@zachary1977information]: 34 members of
a 1970s university club, 78 friendships, and --- famously --- a schism that
split the club in two after a dispute between the instructor (node 1) and the
president (node 34).

```{r}
#| label: fig-karate
#| fig-cap: "Zachary's karate club, coloured by the faction each member joined after the split. Nobody told the layout algorithm about the factions."
#| fig-width: 7
#| fig-height: 5
g <- karate_graph()
A <- as_adj_dense(g)
lab <- V(g)$label

set.seed(4)
coords <- layout_with_fr(g)
par(mar = c(0, 0, 0, 0))
plot(g,
  layout = coords,
  vertex.color = ifelse(lab == "Mr Hi", "#4C72B0", "#DD8452"),
  vertex.label.color = "white", vertex.size = 15,
  vertex.label.cex = 0.7, edge.color = "grey75"
)
legend("topleft", legend = c("Mr Hi", "Officer"), pch = 21,
       pt.bg = c("#4C72B0", "#DD8452"), bty = "n", pt.cex = 1.6)
```

Now suppose you want to do something ordinary with this: cluster the members,
predict who will befriend whom next, or feed the club into a model that also
uses age and attendance. Every one of those tasks wants a data frame with one
row per member and a fixed number of numeric columns.

What do you have instead? A $34 \times 34$ adjacency matrix:

```{r}
A[1:8, 1:8]
```

You could use it directly --- one column per node --- and people do. It fails
in three ways at once:

1.  **It is the wrong width.** The number of features equals the number of
    nodes. Add a member, and every model you fitted has the wrong input shape.
2.  **It is almost all zeros.** Real graphs have average degree in the tens
    while $n$ is in the millions. Here, `r round(100 * mean(A > 0), 1)`% of
    entries are non-zero, and that is a *dense* graph by real standards.
3.  **It has no useful geometry.** Rows 1 and 34 --- the two faction leaders,
    with 16 and 17 friends each --- share
    `r sum(A[1, ] > 0 & A[34, ] > 0)` common neighbours. But
    `sum(A[1,] * A[34,])` treats "friend in common" as the *only* kind of
    similarity. Two nodes three hops apart in the same tight community look
    exactly as unrelated as two nodes in different components.

Point 3 is the real problem, and it does not go away by being clever with
sparse storage. The adjacency matrix records edges. What you want is a notion of
*similarity* --- and in a graph, similarity is a global property that emerges
from paths, not a local one you can read off a single entry.

## The idea

::: {.callout-note appearance="simple"}
## Graph embedding

A **graph embedding** is a map $f: V \to \mathbb{R}^d$ from nodes to short
vectors, chosen so that **geometric closeness in $\mathbb{R}^d$ approximates
structural closeness in $G$**.
:::

That is the entire concept. $d$ is typically 16 to 256 --- small, dense, and
fixed regardless of $n$. The vectors are usually meaningless individually
(dimension 7 does not "mean" anything) and meaningful in relation to each
other.

To see what "good" looks like, here is the karate club embedded into two
dimensions by the very first algorithm we will study. No labels were used:

```{r}
#| label: fig-karate-embed
#| fig-cap: "The karate club embedded in 2D by a Laplacian eigenmap. Colour is the true faction, which the algorithm never saw."
#| fig-width: 7
#| fig-height: 4.5
Z <- spectral_embedding(A, d = 2)

par(mar = c(4, 4, 1, 1))
plot(Z[, 1], Z[, 2], type = "n", xlab = "dimension 1", ylab = "dimension 2")
abline(v = 0, col = "grey85", lty = 2)
text(Z[, 1], Z[, 2], labels = seq_len(nrow(Z)), cex = 0.8,
     col = ifelse(lab == "Mr Hi", "#4C72B0", "#DD8452"))
```

The first dimension alone separates the factions almost perfectly: thresholding
it at zero recovers the true split for 33 of the 34 members. The two leaders,
1 and 34, land on opposite sides of zero. Node 9 --- a genuinely ambiguous
case, who joined the Officer faction despite being better connected to Mr Hi's
group --- comes out ranked 17th of 34 along that axis, dead in the middle,
which is exactly where an honest method should put it.

Note who is *not* at the extremes: the ends of the axis are held by peripheral
members like 17 and 30, not by the two hubs. Laplacian eigenmaps push
loosely-attached nodes outward, because moving them is cheap. That is a real
property of the method, and @sec-spectral explains where it comes from.

Two sentences of arithmetic did that, and the rest of this book is about what
those two sentences can be.

## The encoder–decoder framing

Almost every method in the literature fits one template [@hamilton2020]. It is
worth internalising, because it turns a zoo of algorithms into three choices.

You pick:

1.  An **encoder** $\textsc{enc}(i) = \mathbf{z}_i$, producing the vector for
    node $i$.
2.  A **decoder** $\textsc{dec}(\mathbf{z}_i, \mathbf{z}_j)$, reconstructing
    some quantity from a pair of vectors --- usually a dot product
    $\mathbf{z}_i^\top \mathbf{z}_j$.
3.  A **target similarity** $S_{ij}$: the ground truth the decoder is trying to
    match.

Then you minimise

$$
\mathcal{L} = \sum_{(i,j)} \ell\bigl(\textsc{dec}(\mathbf{z}_i, \mathbf{z}_j),\; S_{ij}\bigr).
$$

Everything is a choice of $S$ and $\ell$:

| Method | Target similarity $S_{ij}$ | Loss |
|---|---|---|
| Laplacian eigenmaps | $A_{ij}$ | weighted squared distance |
| Graph factorisation | $A_{ij}$ | squared error on dot product |
| DeepWalk / node2vec | probability of visiting $j$ from $i$ in a short random walk | cross-entropy (with negative sampling) |
| NetMF | log of that same walk probability, in closed form | squared error |
| GCN / GraphSAGE | task-dependent | task-dependent |

Two things follow immediately.

**The interesting choice is $S$, not the optimiser.** The reason DeepWalk beats
plain adjacency factorisation is not that it uses a neural network --- we will
see in @sec-mf that it barely does. It is that random walks define a
similarity that reaches beyond immediate neighbours.

**Most encoders are just a lookup table.** For everything in @sec-spectral
through @sec-mf, $\textsc{enc}$ is literally "row $i$ of a matrix $Z$", with
$Z$ the free parameter. These are called **shallow** embeddings, and they have
a serious consequence covered next.

## Shallow vs. deep, transductive vs. inductive

If your encoder is a lookup table with $n \times d$ free parameters, then:

-   You cannot embed a node you did not train on. A new member joins the club;
    you have no row for them. The only fix is to retrain.
-   You cannot use node attributes. If members have an age and an attendance
    record, a lookup table has nowhere to put them.
-   Parameters grow linearly with $n$. A hundred million nodes at $d = 128$ is
    51 GB of embedding table.

This is the **transductive** setting: you embed a fixed, known graph. It is
fine for a citation network you analyse once, and useless for a product
catalogue that gains items hourly.

The alternative is an encoder that is a *function of a node's neighbourhood and
features* rather than its identity:

$$
\mathbf{z}_i = f_\theta(\mathbf{x}_i, \{\mathbf{x}_j : j \in \mathcal{N}(i)\}).
$$

Now the parameters $\theta$ are shared across all nodes, the count does not
grow with $n$, and a brand-new node gets an embedding by running $f_\theta$ on
its neighbourhood. That is the **inductive** setting, and it is what graph
neural networks buy you (@sec-gnn).

## The distinction that decides everything

There are two completely different things people mean by "these two nodes are
similar", and choosing the wrong one is the most common way to get a bad
embedding.

**Homophily / proximity.** Nodes are similar if they are close --- same
community, short paths between them. Two friends in the same club. This is what
you want for recommendation, community detection, and most link prediction.

**Structural equivalence / role.** Nodes are similar if they play the same
*role*, regardless of where they sit. Two bridges between communities are
similar even if they are on opposite sides of the graph and share no
neighbours. This is what you want for anomaly detection, or for asking "which
of my users behave like influencers?".

The barbell graph makes the conflict concrete: two cliques joined by a path.

```{r}
#| label: fig-barbell
#| fig-cap: "A barbell. Nodes 1 and 7 (the clique attachment points) are far apart but structurally identical. Nodes 14, 15, 16 are close together and also structurally identical."
#| fig-width: 7
#| fig-height: 3.4
gb <- barbell_graph(clique = 6, path = 5)
set.seed(2)
par(mar = c(0, 0, 0, 0))
plot(gb, layout = layout_with_kk(gb), vertex.size = 20,
     vertex.color = "#4C72B0", vertex.label.color = "white",
     vertex.label.cex = 0.8, edge.color = "grey70")
```

Node 1 and node 7 are the two attachment points. A proximity-based embedding
puts them far apart --- correctly, they are 6 hops away. A role-based embedding
puts them on top of each other --- also correctly, they are indistinguishable
if you relabel the graph.

Neither is wrong. They answer different questions. node2vec (@sec-walks) is
famous precisely because it exposes a knob that interpolates between the two,
and we will turn that knob and watch this graph respond.

## Why bother, concretely

Three places this pays off, to keep the abstraction honest.

**Recommendation.** Users and items form a bipartite graph. Embed both into the
same space; recommend by nearest neighbour. This is roughly what every
large-scale recommender did before, and often still inside, deep models.

**Medical vocabularies.** The OMOP/SNOMED concept hierarchy is a graph of a few
million clinical concepts joined by "is a", "maps to", and anatomical
relations. Embedding it gives you a similarity between diagnoses that respects
clinical structure rather than string edit distance --- so "myocardial
infarction" lands near "acute coronary syndrome" and nowhere near
"myocardial bridge". Those vectors then become the token embeddings for a model
over patient event sequences, which is exactly the hand-off in
@sec-transformers.

**Fraud and abuse.** Accounts, devices, payment instruments, and IPs form a
heterogeneous graph. Rings of collusion are dense subgraphs; embedding turns
"is part of a suspicious cluster" into "is near these other vectors", which a
downstream classifier can use alongside ordinary tabular features.

## What makes an embedding good

We need to be able to answer this, so here is the short version; @sec-eval does
it properly.

-   **Link prediction.** Hide 15% of edges, embed the rest, score every
    candidate pair by $\mathbf{z}_i^\top \mathbf{z}_j$, and check that the
    held-out edges outrank random non-edges. Reported as AUC.
-   **Node classification.** If you have labels, do they become linearly
    separable --- or at least k-NN-separable --- in embedding space?
-   **Do the neighbourhoods look right?** The oldest and most underrated
    method: pick a few nodes, list their nearest neighbours, and see whether a
    domain expert nods.

Here is that last check on the embedding from earlier, using more dimensions:

```{r}
Z8 <- spectral_embedding(A, d = 8)
S <- cosine_sim(Z8)
diag(S) <- -Inf

for (i in c(1, 34, 9)) {
  nb <- order(S[i, ], decreasing = TRUE)[1:5]
  cat(sprintf(
    "node %2d (%-7s): %s\n", i, lab[i],
    paste(sprintf("%d(%s)", nb, substr(lab[nb], 1, 2)), collapse = "  ")
  ))
}
```

Nodes 1 and 34 retrieve neighbours from their own faction. Node 9 --- the
defector --- retrieves a mix, which is the correct answer for a genuinely
in-between node.

Now let us build these things.
