# Spectral embeddings {#sec-spectral}

```{r}
#| label: setup
source("R/graphemb.R")
set.seed(1)
g <- karate_graph()
A <- as_adj_dense(g)
lab <- V(g)$label
```

The oldest family, and the one everything else should be measured against. It
has no hyperparameters worth tuning, no randomness, a closed-form solution, and
in a lot of published comparisons it is within noise of the neural methods.
Start here.

## The objective

We want an embedding where connected nodes end up close. Write that down
literally. For a one-dimensional embedding $\mathbf{f} \in \mathbb{R}^n$,
penalise every edge by the squared gap it has to stretch:

$$
\mathcal{L}(\mathbf{f}) = \tfrac{1}{2}\sum_{i,j} A_{ij}\,(f_i - f_j)^2 .
$$

This is small when every edge connects nodes with similar coordinates. Now the
identity that makes the whole field work: expand the square.

$$
\begin{aligned}
\tfrac{1}{2}\sum_{i,j} A_{ij}(f_i - f_j)^2
&= \tfrac{1}{2}\sum_{i,j} A_{ij}(f_i^2 - 2f_if_j + f_j^2) \\
&= \sum_i d_i f_i^2 - \sum_{i,j} A_{ij} f_i f_j \\
&= \mathbf{f}^\top D \mathbf{f} - \mathbf{f}^\top A \mathbf{f}
 \;=\; \mathbf{f}^\top (D - A) \mathbf{f} .
\end{aligned}
$$

The matrix in the middle,

$$
L = D - A,
$$

is the **graph Laplacian**. The sum-over-edges objective *is* a quadratic form
in $L$. Minimising smoothness over a graph and doing linear algebra with $L$
are the same activity.

Let us confirm that identity numerically rather than take it on faith:

```{r}
L <- graph_laplacian(A, "comb")
f <- rnorm(nrow(A))

edge_sum <- 0.5 * sum(outer(f, f, "-")^2 * A)
quad_form <- as.numeric(t(f) %*% L %*% f)

c(edge_sum = edge_sum, quadratic_form = quad_form)
```

Identical, as promised.

## Why we need constraints

Minimising $\mathbf{f}^\top L \mathbf{f}$ without constraints gives
$\mathbf{f} = \mathbf{0}$: perfectly smooth, perfectly useless. Two constraints
fix this.

**Scale.** Require $\mathbf{f}^\top D \mathbf{f} = 1$ so the embedding cannot
collapse. (Using $D$ rather than $I$ weights nodes by degree, which turns out to
be the right choice --- see below.)

**Non-triviality.** The constant vector $\mathbf{f} = \mathbf{1}$ has
$L\mathbf{1} = \mathbf{0}$: it costs nothing because it stretches no edge. It
is the trivial solution and we must exclude it, so require
$\mathbf{f}^\top D \mathbf{1} = 0$.

With those, the minimiser is the eigenvector of the generalised problem

$$
L \mathbf{f} = \lambda D \mathbf{f}
$$

with the **smallest non-zero** eigenvalue. For a $d$-dimensional embedding, take
the next $d$ eigenvectors. This is the **Laplacian eigenmap** of
@belkin2003laplacian.

::: {.callout-tip appearance="simple"}
## The one-line summary

A spectral embedding is the set of **smoothest non-constant functions on the
graph** --- the coordinates that change least as you walk along an edge.
:::

## Three Laplacians

You will meet three, and mixing them up is a common source of confusion.

| Name | Definition | Note |
|---|---|---|
| Combinatorial | $L = D - A$ | Unnormalised. Sensitive to degree; hubs dominate |
| Symmetric | $L_{\mathrm{sym}} = I - D^{-1/2} A D^{-1/2}$ | Symmetric, eigenvalues in $[0, 2]$. What you compute with |
| Random-walk | $L_{\mathrm{rw}} = I - D^{-1} A$ | $D^{-1}A$ is the walk transition matrix. What you interpret with |

The last two share eigenvalues, and their eigenvectors are related by
$\mathbf{f}_{\mathrm{rw}} = D^{-1/2}\mathbf{f}_{\mathrm{sym}}$. So the standard
recipe is: compute with $L_{\mathrm{sym}}$ because it is symmetric and
eigensolvers like symmetric matrices, then multiply by $D^{-1/2}$ to get the
vectors you actually want. Here is that, in full:

```{r}
graph_laplacian
```

## The implementation

The whole algorithm:

```{r}
spectral_embedding
```

Four things worth pointing at:

-   `eigen(..., symmetric = TRUE)` returns eigenvalues in **decreasing** order,
    so we reverse to get the smallest.
-   `drop_first` discards the trivial constant eigenvector. Forgetting this is
    the single most common bug in hand-rolled spectral code: your first
    embedding dimension comes out as a constant and every downstream model
    silently ignores it.
-   The `type == "sym"` branch does the $D^{-1/2}$ correction described above.
-   The eigenvalues come back as an attribute, because they are diagnostic.

## The eigenvalues tell you the community structure

$\lambda_2$, the smallest non-zero eigenvalue, is the **Fiedler value** or
algebraic connectivity. It measures how hard the graph is to cut in two: near
zero means there is a good bottleneck, and its eigenvector tells you where to
cut.

More usefully, a *gap* in the eigenvalue sequence tells you how many
communities there are. If the graph has $k$ well-separated clusters, the first
$k$ eigenvalues sit near zero and then there is a jump.

```{r}
#| label: fig-eigengap
#| fig-cap: "Eigenvalues of the symmetric Laplacian. Left: the karate club, with one clear near-zero eigenvalue after the trivial one, matching its two factions. Right: a planted-partition graph with three communities, showing three."
#| fig-width: 8
#| fig-height: 3.6
g3 <- sbm_graph(n_per = 40, k = 3, p_in = 0.14, p_out = 0.01, seed = 42)
A3 <- as_adj_dense(g3)

ev_karate <- sort(eigen(graph_laplacian(A, "sym"), symmetric = TRUE)$values)
ev_sbm <- sort(eigen(graph_laplacian(A3, "sym"), symmetric = TRUE)$values)

par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(ev_karate[1:10], type = "b", pch = 19, col = "#4C72B0",
     xlab = "index", ylab = expression(lambda), main = "karate club (2 factions)")
abline(h = 0, col = "grey85")
plot(ev_sbm[1:10], type = "b", pch = 19, col = "#DD8452",
     xlab = "index", ylab = expression(lambda), main = "planted partition (3 blocks)")
abline(h = 0, col = "grey85")
```

```{r}
round(ev_sbm[1:6], 4)
```

Three eigenvalues near zero, then a jump. The graph told us how many
communities it has without being asked. This is the basis of **spectral
clustering**: embed with the first $k$ eigenvectors, then run k-means.

```{r}
set.seed(1)
Z3 <- spectral_embedding(A3, d = 3)
km <- kmeans(normalize_rows(Z3), centers = 3, nstart = 20)
table(cluster = km$cluster, truth = V(g3)$label)
```

A near-perfect recovery of the planted communities from three eigenvectors and
k-means.

## Where the axis extremes come from

In @sec-idea we noticed that the karate club's most extreme nodes along
dimension 1 were peripheral members, not the two hubs. Now we can say why.

The constraint is $\mathbf{f}^\top D \mathbf{f} = \sum_i d_i f_i^2 = 1$. A node
with large degree pays a large price for a large coordinate, so the optimiser
keeps hubs near the origin and spends its budget moving low-degree nodes. The
degree weighting is doing exactly what it was designed to do; it just means
"extreme along the Fiedler vector" reads as "loosely attached", not "important".

```{r}
d1 <- spectral_embedding(A, d = 1)[, 1]
deg <- rowSums(A)
data.frame(
  node = seq_along(d1), degree = deg, dim1 = round(d1, 3)
)[order(abs(d1), decreasing = TRUE)[1:6], ]
```

The six most extreme nodes have degrees of
`r paste(sort(deg[order(abs(d1), decreasing = TRUE)[1:6]]), collapse = ", ")`,
against a graph median of `r median(deg)` and a maximum of `r max(deg)`. The
axis is ranking attachment strength, not importance.

## It recovers real geometry

The sharpest demonstration: build a graph from points in the plane, joining
pairs that are close. Throw the coordinates away, keep only the edges, and ask
the embedding to reconstruct them.

```{r}
#| label: fig-geo
#| fig-cap: "Left: a random geometric graph drawn at its true coordinates. Right: the 2D spectral embedding, which never saw those coordinates. Colour is the true x-position."
#| fig-width: 8.5
#| fig-height: 4.2
gg <- geo_graph(n = 300, radius = 0.115, seed = 7)
Ag <- as_adj_dense(gg)
xy <- cbind(V(gg)$x, V(gg)$y)
Zg <- spectral_embedding(Ag, d = 2)

pal <- colorRampPalette(c("#4C72B0", "#F2F2F2", "#DD8452"))(100)
col <- pal[cut(xy[, 1], 100, labels = FALSE)]

par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(xy, pch = 19, cex = 0.7, col = col, xlab = "x", ylab = "y",
     main = "true coordinates")
plot(Zg, pch = 19, cex = 0.7, col = col, xlab = "dim 1", ylab = "dim 2",
     main = "spectral embedding")
```

The embedding is stretched and rotated --- eigenvectors carry no canonical
orientation or scale --- so comparing axis to axis understates the match:

```{r}
round(abs(cor(Zg, xy)), 3)
```

Every entry is middling, because embedding dimension 1 is a *mixture* of true
$x$ and $y$. The rotation-invariant question is whether the two-dimensional
subspaces agree, which is what canonical correlation measures:

```{r}
round(cancor(Zg, xy)$cor, 3)
```

`r round(cancor(Zg, xy)$cor[1], 2)` and `r round(cancor(Zg, xy)$cor[2], 2)`.
The embedding recovered the latent plane the graph was generated from, given
nothing but a list of edges.

## Cost, and when it stops working

`eigen()` is $O(n^3)$ and needs the dense matrix in memory. At $n = 10{,}000$
that is 800 MB and a few minutes; at $n = 10^6$ it is impossible.

The fix is that you never want all $n$ eigenvectors, only the first few, and
iterative solvers get those from sparse matrix–vector products alone. In R,
**RSpectra** is the right tool:

```{r}
#| eval: false
library(RSpectra)
library(Matrix)

A_sp <- as(A, "dgCMatrix")
d <- rowSums(A_sp)
L_sym <- Diagonal(nrow(A_sp)) - Diagonal(x = 1 / sqrt(d)) %*% A_sp %*% Diagonal(x = 1 / sqrt(d))

# 'SM' = smallest magnitude; ask for d+1 and drop the trivial one
res <- eigs_sym(L_sym, k = 9, which = "SM")
Z <- res$vectors[, ncol(res$vectors):1][, -1] / sqrt(d)
```

In practice, shift-and-invert (`sigma = 0`) converges far faster than `"SM"`
for Laplacians, because the smallest eigenvalues are tightly clustered.

## Strengths and limits

**Good:** deterministic, no tuning, fast on moderate graphs, gives you the
number of communities for free, and has a genuine theory --- it is a relaxation
of normalised cut, with bounds relating $\lambda_2$ to the quality of the best
partition.

**Bad:**

-   *Global and rigid.* Every dimension is a global function of the whole
    graph. You cannot say "embed this subgraph well and don't worry about the
    rest".
-   *Purely proximity-based.* It has no way to express structural roles.
-   *Transductive.* Add a node and you re-run the eigendecomposition.
-   *No node features.*
-   *Degree-dominated extremes*, as we saw.

The next chapter attacks the first two by replacing "smooth over edges" with
"co-occurs in a random walk", which reaches beyond immediate neighbours and
comes with a knob for proximity vs. role.
