# DeepWalk is a matrix factorisation {#sec-mf}

```{r}
#| label: setup
source("R/graphemb.R")
set.seed(1)
g <- karate_graph()
A <- as_adj_dense(g)
lab <- V(g)$label
```

This is my favourite result in the area, and the one I would keep if the talk
were three minutes instead of seven.

We just spent a chapter sampling random walks and running stochastic gradient
descent to learn embeddings. It worked. It was also, in a precise sense,
unnecessary: the thing that procedure converges to has a closed form, and you
can compute it with an SVD in milliseconds.

## Part 1: skip-gram is factorising a PMI matrix

@levy2014neural proved this for word2vec. Fix the model: we are learning
$\mathbf{w}_u$ and $\mathbf{c}_v$ and scoring pairs by
$s_{uv} = \mathbf{w}_u^\top \mathbf{c}_v$. Forget the low-rank constraint for a
moment and ask: if each $s_{uv}$ could take any value independently, what value
minimises the SGNS loss?

The expected loss contributed by the pair $(u,v)$, over observed
co-occurrences and sampled negatives, is

$$
\mathcal{L}_{uv} = -\,\#(u,v)\,\log\sigma(s_{uv})
\;-\; K \cdot \#(u)\,\frac{\#(v)}{|\mathcal{D}|}\,\log\sigma(-s_{uv}),
$$

where $\#(u,v)$ counts co-occurrences, $\#(u)$ and $\#(v)$ are marginals, and
$|\mathcal{D}|$ is the total number of pairs. Differentiate with respect to
$s_{uv}$, set to zero, and using $\sigma'(x) = \sigma(x)\sigma(-x)$ you get

$$
s_{uv} = \log\!\left(\frac{\#(u,v)\,|\mathcal{D}|}{\#(u)\,\#(v)}\right) - \log K .
$$

The first term is the **pointwise mutual information** of the pair. So the
optimal score is

$$
\mathbf{w}_u^\top \mathbf{c}_v = \mathrm{PMI}(u,v) - \log K .
$$

Skip-gram with negative sampling is performing an implicit low-rank
factorisation of a shifted PMI matrix. The neural framing is a way of computing
it by sampling; it is not where the representational power comes from.

We can check the claim directly. Fit SGNS, then compare the learned dot
products against the empirical shifted PMI computed from the very same pairs:

```{r}
#| label: verify-pmi
W <- random_walks(A, num_walks = 40, walk_length = 40, seed = 1)
P <- walks_to_pairs(W, window = 5, seed = 1)
n <- nrow(A)
K <- 5

fit <- skipgram_sgns(P, n_nodes = n, dim = 32, epochs = 12, neg = K, seed = 1)

# empirical co-occurrence counts and marginals from the same corpus
Cnt <- matrix(0, n, n)
tab <- table(P[, 1], P[, 2])
Cnt[cbind(as.integer(rownames(tab))[row(tab)],
          as.integer(colnames(tab))[col(tab)])] <- as.vector(tab)

Dtot <- sum(Cnt)
pmi <- log((Cnt * Dtot) / outer(rowSums(Cnt), colSums(Cnt))) - log(K)

learned <- fit$W %*% t(fit$C)
keep <- is.finite(pmi) & Cnt > 20      # ignore pairs seen too rarely to estimate

cor(learned[keep], pmi[keep])
```

```{r}
#| label: fig-pmi
#| fig-cap: "Learned dot products against empirical shifted PMI, for node pairs observed at least 20 times. SGNS is fitting the PMI matrix, up to the error of a rank-32 approximation."
#| fig-width: 6.5
#| fig-height: 4.5
par(mar = c(4.2, 4.2, 1, 1))
plot(pmi[keep], learned[keep], pch = 19, cex = 0.35,
     col = adjustcolor("#4C72B0", 0.4),
     xlab = "empirical shifted PMI", ylab = expression(w[u] %*% c[v]))
abline(0, 1, col = "#DD8452", lwd = 2)
```

A correlation of `r round(cor(learned[keep], pmi[keep]), 2)` with the identity
line drawn on top. It is not exact --- rank 32 cannot reproduce a full-rank
matrix, and the rarer pairs are noisily estimated --- but the relationship is
unmistakable. The neural network is fitting the statistic the theory says it
should.

## Part 2: the co-occurrence counts have a closed form

Now the graph-specific half, from @qiu2018network.

Why sample walks at all? A random walk's transition matrix is
$P = D^{-1}A$. The probability of being at $j$ exactly $r$ steps after starting
at $i$ is $[P^r]_{ij}$ --- no sampling required. Averaging over window offsets
$r = 1, \dots, T$ and folding in the marginals gives, in expectation over an
infinite corpus of walks,

$$
M = \frac{\mathrm{vol}(G)}{bT}\left(\sum_{r=1}^{T} P^r\right) D^{-1},
\qquad \mathrm{vol}(G) = \sum_{ij} A_{ij},
$$

and DeepWalk's target is $\log \max(M, 1)$ --- the truncation replacing
$\log$ of anything below 1 with zero, which is the same "positive PMI" trick
used throughout the word-vector literature.

This is **NetMF**. Two functions:

```{r}
netmf_matrix
```

```{r}
netmf
```

That is the whole algorithm. No walks, no SGD, no epochs, no learning rate.

## Do they agree?

The claim is that sampled DeepWalk and closed-form NetMF estimate the same
object, so the sampled version should converge to the analytic one as we draw
more walks.

One subtlety has to be dealt with first, and it is the kind of thing that
quietly ruins a comparison. The formula above weights every walk offset
$r = 1, \dots, T$ **equally**. Word2vec's shrinking window does not: an offset
$r$ pair survives with probability $(T - r + 1)/T$, so near context is
up-weighted. That is a different estimand, and it shows up as a bias that no
amount of extra sampling removes. For the comparison we therefore switch the
shrinking off, which is what `shrink = FALSE` is for.

```{r}
#| label: converge
M_exact <- netmf_matrix(A, T_window = 5, b = 1, truncate = FALSE)

empirical_pmi <- function(r, shrink) {
  Wr <- random_walks(A, num_walks = r, walk_length = 40, seed = 2)
  Pr <- walks_to_pairs(Wr, window = 5, shrink = shrink, seed = 2)
  Cr <- matrix(0, n, n)
  tb <- table(Pr[, 1], Pr[, 2])
  Cr[cbind(as.integer(rownames(tb))[row(tb)],
           as.integer(colnames(tb))[col(tb)])] <- as.vector(tb)
  log((Cr * sum(Cr)) / outer(rowSums(Cr), colSums(Cr)))
}

# Freeze the set of pairs we score on, so the comparison is like-for-like:
# well-observed, and seen at least once in the sparsest run of either setting
# (an unobserved pair has empirical PMI of -Inf and would poison the mean).
keep2 <- Cnt > 20 & is.finite(M_exact) &
  empirical_pmi(5, FALSE) > -Inf & empirical_pmi(5, TRUE) > -Inf

walks <- c(5, 20, 80, 320)
res <- sapply(walks, function(r) {
  pm <- empirical_pmi(r, shrink = FALSE)
  c(rmse = sqrt(mean((M_exact[keep2] - pm[keep2])^2)),
    cor  = cor(M_exact[keep2], pm[keep2]))
})
colnames(res) <- paste0(walks, " walks/node")
round(res, 4)
```

The error falls and the correlation climbs to `r round(max(res["cor", ]), 3)`.
The walks really were estimating this matrix, and the gap is Monte Carlo error.

Run the same sweep with the shrinking window and the bias is plain:

```{r}
round(sapply(walks, function(r) {
  pm <- empirical_pmi(r, shrink = TRUE)
  sqrt(mean((M_exact[keep2] - pm[keep2])^2))
}), 4)
```

It improves, then stalls around 0.26 --- three times the uniform-window error,
and still falling far more slowly. It is converging, just to a different
matrix. If you want NetMF and your DeepWalk implementation to agree, they have
to agree about the window first.

The residual floor in the uniform case (about 0.08) is not sampling noise
either: our walks are 40 steps long and start from a uniform distribution over
nodes, while the derivation assumes the walk is in its stationary distribution.
Short walks under-represent long-range co-occurrence. Lengthen the walks and it
shrinks further.

## Which is better in practice?

```{r}
#| label: compare
t_dw <- system.time(Z_dw <- deepwalk(A, dim = 32, epochs = 5, seed = 1))[["elapsed"]]
t_nm <- system.time(Z_nm <- netmf(A, dim = 32, T_window = 5))[["elapsed"]]
t_sp <- system.time(Z_sp <- spectral_embedding(A, d = 8))[["elapsed"]]

data.frame(
  method = c("DeepWalk (sampled + SGD)", "NetMF (closed form)", "Spectral"),
  seconds = round(c(t_dw, t_nm, t_sp), 3),
  knn_accuracy = round(c(knn_accuracy(Z_dw, lab, 5),
                         knn_accuracy(Z_nm, lab, 5),
                         knn_accuracy(Z_sp, lab, 5)), 3)
)
```

DeepWalk and NetMF tie at perfect faction recovery, spectral is a fraction
behind, and NetMF gets there roughly a thousand times faster than the sampled
version --- microseconds against seconds.

## The point, and the catch

::: {.callout-important appearance="simple"}
## What this buys you

Before you reach for a "graph representation learning" library, try an SVD of
$\log \max(M, 1)$. It is deterministic, has one meaningful hyperparameter
($T$), gives you a reproducible answer, and on many benchmarks matches the
sampled methods. When it is not good enough, you have a strong baseline that
tells you how much the complexity is buying.
:::

The catch is right there in the code: `netmf_matrix` builds a **dense $n \times
n$ matrix** and multiplies it $T$ times. That is $O(n^3)$ time and $O(n^2)$
memory. At $n = 10^5$ the matrix alone is 80 GB. DeepWalk's sampling is
"unnecessary" only in the sense that it approximates something exact --- the
approximation is the entire reason it scales.

Qiu et al. addressed this with NetMF's large-window variant, which never forms
$M$: eigendecompose $D^{-1/2} A D^{-1/2}$, apply the polynomial
$\sum_r \lambda^r$ to the eigenvalues, and reconstruct in low rank. Their
follow-up, NetSMF, sparsifies the matrix instead. Both keep the closed-form
character while avoiding the dense product.

So the honest summary:

| Graph size | Reach for |
|---|---|
| Up to ~20k nodes | NetMF, exactly as written here |
| 20k – 1M | Sparse spectral (RSpectra) or NetMF's large-window variant |
| Beyond that | Sampled walks + SGNS, or a GNN |

## The unification

Step back and every method so far collapses into one recipe.

$$
\underbrace{Z}_{\text{embedding}}
\;\approx\;
\text{low-rank factorisation of}
\;\underbrace{f(A)}_{\substack{\text{a matrix of} \\ \text{node similarities}}}
$$

| Method | $f(A)$ |
|---|---|
| Spectral | $I - D^{-1/2}AD^{-1/2}$ (bottom eigenvectors) |
| Graph factorisation | $A$ |
| DeepWalk / NetMF | $\log\max\!\big(\tfrac{\mathrm{vol}(G)}{bT}(\sum_r P^r)D^{-1},\,1\big)$ |
| node2vec | the same, with $P$ replaced by a second-order transition operator |
| LINE [@tang2015line] | the $T = 1$ case of NetMF |

They differ in one place: **which matrix**. Everything else --- SGD versus SVD,
neural versus not --- is implementation.

That framing also shows the ceiling. Every row of that table is a function of
$A$ alone. None of them can use the fact that a node is a 34-year-old who
joined in March, and none can embed a node that was not in the matrix when you
factorised it. Breaking through that is the next chapter.
