# Graph neural networks {#sec-gnn}

```{r}
#| label: setup
source("R/graphemb.R")
set.seed(1)
g <- karate_graph()
A <- as_adj_dense(g)
lab <- V(g)$label
```

Everything so far factorises a matrix built from $A$. That ceiling has two
consequences we keep bumping into: node attributes cannot be used, and a node
absent at training time cannot be embedded at all.

Graph neural networks break both by replacing the lookup table with a
**function of a node's neighbourhood**.

## Message passing

One layer, in three steps. For every node $i$:

1.  **Message.** Each neighbour $j \in \mathcal{N}(i)$ emits a message computed
    from its current representation $\mathbf{h}_j$.
2.  **Aggregate.** Combine the messages with a permutation-invariant operator
    --- sum, mean, or max. It *must* be permutation-invariant, because a
    node's neighbours have no canonical order.
3.  **Update.** Combine the aggregate with $\mathbf{h}_i$ to get
    $\mathbf{h}_i'$.

$$
\mathbf{h}_i^{(l+1)} = \phi\left(\mathbf{h}_i^{(l)},\;
\bigoplus_{j \in \mathcal{N}(i)} \psi\!\left(\mathbf{h}_j^{(l)}\right)\right)
$$

Stack $L$ layers and information travels $L$ hops. That is the whole framework;
GCN, GraphSAGE, GAT and the rest are choices of $\psi$, $\bigoplus$ and $\phi$.

## The GCN layer

@kipf2017semi made the simplest choice that works, and it is still the default
baseline. Written for all nodes at once:

$$
H^{(l+1)} = \sigma\!\left(\hat{A}\, H^{(l)}\, W^{(l)}\right),
\qquad
\hat{A} = \tilde{D}^{-1/2}(A + I)\tilde{D}^{-1/2},
\quad \tilde{D} = \mathrm{diag}(A + I)\mathbf{1}.
$$

Three ideas compressed into one line:

-   **$A + I$** adds a self-loop, so a node keeps its own features instead of
    being replaced by its neighbourhood.
-   **$\tilde{D}^{-1/2} \cdot \tilde{D}^{-1/2}$** normalises symmetrically.
    Without it, high-degree nodes produce huge activations and training
    diverges. This is not the same as the row-normalisation $\tilde D^{-1}(A+I)$
    used by a mean-aggregator: the symmetric version down-weights messages
    *from* high-degree neighbours as well as *into* high-degree nodes.
-   **$W^{(l)}$** is shared by every node. That is what makes the model
    inductive: the parameter count depends on feature width, not on $n$.

```{r}
gcn_normalize
```

```{r}
gcn_forward
```

Note what $\hat A H$ actually does: it replaces each node's feature vector with
a weighted average of its neighbourhood. **A GCN layer is a learned smoother on
the graph.** That framing explains most of its behaviour, good and bad.

## Message passing works before any training

A striking property: even with *random* weights, a GCN produces useful
embeddings, because the smoothing is doing the work. Here is a two-layer GCN
with random weights on the karate club, given no features at all --- the input
is the identity matrix, i.e. a one-hot per node.

```{r}
#| label: fig-gcn-random
#| fig-cap: "A two-layer GCN with random, untrained weights, applied to one-hot node inputs. The factions separate anyway: neighbourhood averaging is doing the work."
#| fig-width: 7
#| fig-height: 4.5
n <- nrow(A)
Ws <- gcn_random_weights(c(n, 32, 2), seed = 3)
H <- gcn_forward(A, diag(n), Ws)

par(mar = c(4, 4, 1, 1))
plot(H, type = "n", xlab = "unit 1", ylab = "unit 2")
text(H[, 1], H[, 2], labels = seq_len(n), cex = 0.8,
     col = ifelse(lab == "Mr Hi", "#4C72B0", "#DD8452"))
```

```{r}
knn_accuracy(H, lab, k = 5)
```

A randomly initialised network recovers most of the faction split. But one seed
proves nothing, so run several:

```{r}
sapply(1:8, function(s) {
  Wr <- gcn_random_weights(c(n, 32, 2), seed = s)
  round(knn_accuracy(gcn_forward(A, diag(n), Wr), lab, k = 5), 3)
})
```

Mostly 0.88–0.94, with the occasional bad draw down near chance. Two lessons,
and the second is the important one:

-   A good chunk of GNN performance comes from the **graph operator**, not from
    learning. When you read a GNN paper, look for an untrained baseline; when
    you write one, include it.
-   Projecting to two dimensions with random weights is lossy, and sometimes
    the draw is unlucky. Single-seed results from randomly initialised networks
    are not results.

## Over-smoothing: the failure mode

If one layer averages over neighbours, what do twenty do? They average over the
whole graph. Every node converges to the same vector and all information is
destroyed. This is **over-smoothing**, and it is why GCNs are almost always
2–3 layers deep --- unusually shallow for a neural network.

We can watch it happen. Measure how distinguishable the representations remain
after each layer:

```{r}
#| label: oversmooth
X <- diag(n)
Ahat <- gcn_normalize(A)

spread <- numeric(12)
Hl <- X
for (l in 1:12) {
  Hl <- Ahat %*% Hl
  Hn <- normalize_rows(Hl)
  # mean pairwise cosine distance: 1 means all directions differ, 0 means identical
  spread[l] <- 1 - mean(Hn %*% t(Hn))
}
round(spread, 4)
```

```{r}
#| label: fig-oversmooth
#| fig-cap: "Representation diversity after repeated neighbourhood averaging, on a log scale. The straight-line decay is the signature of a power iteration."
#| fig-width: 6.5
#| fig-height: 4
par(mar = c(4.2, 4.2, 1, 1))
plot(spread, type = "b", pch = 19, col = "#4C72B0", log = "y",
     xlab = "propagation steps", ylab = "mean pairwise cosine distance")
abline(v = 3, lty = 2, col = "grey60")
text(3, max(spread) / 2, " typical GCN depth", pos = 4, cex = 0.8, col = "grey40")
```

Diversity falls from `r round(spread[1], 2)` to `r round(spread[12], 3)` --- a
factor of `r round(spread[1] / spread[12])` --- and the log-scale plot is close
to a straight line, which is the tell: the decay is geometric, not gradual.

The reason is that repeated multiplication by $\hat A$ is a power iteration.
Everything converges towards $\hat A$'s top eigenvector at a rate set by the
gap to the second, so a deep GCN does not gently blur its representations, it
collapses them exponentially fast. Twelve layers here would leave every node
holding essentially the same vector, scaled.

## GraphSAGE and the inductive payoff

@hamilton2017inductive made the inductive property explicit and practical.
GraphSAGE differs from GCN in two ways:

$$
\mathbf{h}_i^{(l+1)} = \sigma\!\left(W^{(l)} \cdot
\mathrm{concat}\!\left(\mathbf{h}_i^{(l)},\;
\mathrm{agg}\{\mathbf{h}_j^{(l)} : j \in \mathcal{N}_s(i)\}\right)\right)
$$

-   **Concatenate rather than mix.** A node's own representation is kept in
    separate coordinates from its neighbourhood aggregate, so the layer can
    tell "me" from "around me".
-   **Sample a fixed-size neighbourhood** $\mathcal{N}_s(i)$. This bounds the
    cost per node regardless of degree, which is what makes minibatching over a
    billion-edge graph feasible.

The payoff is the thing shallow embeddings cannot do at all: embed a node that
did not exist at training time. Since the weights are shared, this is just a
forward pass.

```{r}
#| label: inductive
# train-time graph: drop node 34 entirely
keep <- setdiff(seq_len(n), 34)
g_sub <- induced_subgraph(g, keep)
A_sub <- as_adj_dense(g_sub)

# node features that exist for any node, old or new: simple structural stats
node_features <- function(gr) {
  cbind(deg = log1p(degree(gr)),
        clust = transitivity(gr, type = "local", isolates = "zero"),
        core = coreness(gr))
}

X_sub <- scale(node_features(g_sub))
Ws2 <- gcn_random_weights(c(ncol(X_sub), 16, 2), seed = 5)
H_sub <- gcn_forward(A_sub, X_sub, Ws2)

# now node 34 arrives. Same weights, no retraining -- just run the forward pass
# on the full graph, using the training set's scaling.
X_full <- scale(node_features(g),
                center = attr(X_sub, "scaled:center"),
                scale = attr(X_sub, "scaled:scale"))
H_full <- gcn_forward(A, X_full, Ws2)

round(H_full[34, ], 3)
```

Node 34 got an embedding without any parameter ever being refitted. Compare
with a shallow method: `deepwalk()` returns a matrix with `nrow(A_sub)` rows,
and there is simply no row for a node it never saw.

The embeddings for the pre-existing nodes also stay broadly stable, which is
what you need if downstream models are already using them:

```{r}
round(cor(H_sub[, 1], H_full[keep, 1]), 3)
```

A correlation of `r round(cor(H_sub[, 1], H_full[keep, 1]), 2)`. Not identical
--- node 34 has degree 17, so its arrival genuinely changes a third of the
graph's neighbourhoods and the smoother propagates that --- but consistent
enough that downstream models built on the old embeddings do not fall over.

## Graph attention

@velickovic2018graph replaced fixed normalisation weights with learned ones.
Instead of weighting neighbour $j$ by $1/\sqrt{d_i d_j}$, compute

$$
\alpha_{ij} = \frac{\exp\!\big(e_{ij}\big)}{\sum_{k \in \mathcal{N}(i)} \exp(e_{ik})},
\qquad
e_{ij} = \mathrm{LeakyReLU}\!\left(\mathbf{a}^\top [W\mathbf{h}_i \,\|\, W\mathbf{h}_j]\right),
$$

then aggregate with $\sum_j \alpha_{ij} W \mathbf{h}_j$.

This matters for the next chapter, so it is worth stating precisely what
changed: **the aggregation weights now depend on the content of the two
endpoints**, not only on the graph's shape. A node can learn to listen to some
neighbours and ignore others.

And now notice the shape of that expression. A softmax over a set, weighting a
sum of transformed values, with weights computed from pairs of items --- that
is self-attention, restricted to the graph's edges. GAT is a transformer whose
attention mask is the adjacency matrix.

Which raises the obvious question. What if you removed the mask?

## Where GNNs stand

**Use one when** you have node features that matter, the graph changes, or the
task is genuinely semi-supervised and you want to train end-to-end.

**Do not bother when** you have no features and a fixed graph. On plain
structural tasks, a well-tuned NetMF or node2vec is a hard baseline to beat,
and it has no training loop to debug.

**Be aware:**

-   *Over-smoothing* caps depth at 2–3 layers, so the receptive field is small.
    A node cannot be influenced by anything more than $L$ hops away.
-   *Limited expressive power.* Standard message passing is at most as
    discriminative as the 1-Weisfeiler–Lehman graph isomorphism test
    [@xu2019powerful] --- there
    are non-isomorphic graphs it provably cannot tell apart, including some
    simple regular graphs.
-   *Neighbourhood explosion.* Without sampling, an $L$-layer model needs the
    $L$-hop neighbourhood of every node in a batch, which on a graph with hubs
    is most of the graph.

**In R:** there is no mature native GNN stack. The practical routes are the
`torch` package (a full libtorch binding, no Python required) for hand-rolled
layers, or `reticulate` onto PyTorch Geometric or DGL when you want the
optimised kernels and standard datasets.

The next chapter takes the observation that ended the attention section --- that
a GAT is masked self-attention --- and runs it in the other direction.
