# Graph embeddings in a transformer {#sec-transformers}

```{r}
#| label: setup
source("R/graphemb.R")
set.seed(1)
```

## The problem with attention

Self-attention [@vaswani2017attention] computes, for a set of $n$ input vectors
stacked as $X$:

$$
\mathrm{Attn}(X) = \mathrm{softmax}\!\left(\frac{QK^\top}{\sqrt{d_h}}\right)V,
\qquad Q = XW_Q,\; K = XW_K,\; V = XW_V .
$$

Look for where position enters. It does not. Permute the rows of $X$ and the
output rows permute identically: attention is a function on **sets**, not
sequences. That is a feature --- it is why transformers handle long-range
dependencies so well --- and it is why every transformer needs positional
encodings bolted on. Text uses sine waves or learned position vectors because
text has an obvious order.

A graph does not. There is no node 1, node 2, node 3 --- only structure. So:

> How do you tell a permutation-invariant architecture about a structure that
> has no canonical ordering?

That question is exactly what graph embeddings answer, and it is the reason
this chapter exists.

Here is the layer we will use, written out. It is short enough to read:

```{r}
mha
```

```{r}
softmax_rows
```

## Three places structure can enter

There are exactly three, and they are not alternatives --- production models use
two or three at once.

| | Where | What goes in | Cost |
|---|---|---|---|
| **1** | Input embeddings | A pretrained vector per token/entity | Free at train time |
| **2** | Positional encodings | A structural descriptor per node, concatenated or added to $X$ | Cheap preprocessing |
| **3** | Attention bias | A term added to the attention logits, computed from the pair $(i,j)$ | $O(n^2)$ memory |

### 1. Graph embeddings as input vectors

The simplest and by far the most common. You have a transformer over sequences
of *entities*, and those entities also sit in a graph. Embed the graph offline,
use the vectors to initialise the token embedding table, fine-tune or freeze.

The clinical case makes it concrete. Patient records are sequences of coded
events --- diagnoses, drugs, procedures --- which is exactly what a transformer
wants. But the codes also form a graph: the OMOP/SNOMED vocabulary, with "is
a", "maps to", and anatomical relations linking a few million concepts.

Train a randomly initialised embedding table and the model must learn that
*acute myocardial infarction* and *acute coronary syndrome* are related, from
co-occurrence alone, for every one of a million codes --- most of which appear a
handful of times. Initialise from a graph embedding of the vocabulary hierarchy
and that relatedness is there on step zero. The rare codes benefit most,
because their vectors are determined by their position in the hierarchy rather
than by data you do not have.

This is the highest-value, lowest-effort option, and it needs no changes to the
transformer at all.

### 2. Structural positional encodings

Now the case where the graph *is* the input: the tokens are nodes, and we need
to tell attention where each one sits.

**Laplacian positional encoding** [@dwivedi2021generalization] uses the $k$
smallest non-trivial eigenvectors of $L_{\mathrm{sym}}$ --- exactly the
spectral embedding of @sec-spectral --- as a per-node feature vector. It is the
graph analogue of sinusoidal position encodings, and the analogy is precise:
sine waves are the eigenvectors of the Laplacian of a path graph.

```{r}
lap_pe
```

**Random-walk structural encoding** (RWSE, used in GraphGPS by
@rampasek2022recipe) instead records return probabilities:
$[\,P^1_{ii}, P^2_{ii}, \dots, P^K_{ii}\,]$ --- how likely a walk from node $i$
is to be back at $i$ after $k$ steps.

```{r}
rwse
```

The two capture genuinely different things:

```{r}
#| label: pe-compare
g <- karate_graph()
A <- as_adj_dense(g)

pe <- lap_pe(A, k = 4)
rw <- rwse(A, ks = 1:6)

round(head(data.frame(
  node = 1:6, degree = rowSums(A)[1:6],
  lap1 = pe[1:6, 1], lap2 = pe[1:6, 2],
  rw2 = rw[1:6, 2], rw3 = rw[1:6, 3], rw4 = rw[1:6, 4]
)), 3)
```

Column `rw3` is the probability of returning in exactly three steps, which
requires a triangle --- so it senses local clustering. But it is not a
*normalised* clustering coefficient, and the difference trips people up:

```{r}
round(c(
  `vs triangle count` = cor(rw[, 3], count_triangles(g)),
  `vs degree` = cor(rw[, 3], degree(g)),
  `vs transitivity` = cor(rw[, 3], transitivity(g, type = "local", isolates = "zero"))
), 3)
```

Positively correlated with raw triangle count, and *negatively* with the
degree-normalised transitivity. The reason is in the arithmetic: returning in
three steps has probability
$\sum_{a,b} A_{ia}A_{ab}A_{bi} / (d_i d_a d_b)$, so it is discounted not only
by your own degree but by your neighbours'. Look at six nodes that all have
degree 4:

```{r}
i4 <- which(degree(g) == 4)
data.frame(
  node = i4,
  triangles = count_triangles(g)[i4],
  transitivity = round(transitivity(g, type = "local", isolates = "zero")[i4], 3),
  rw3 = round(rwse(A, 1:3)[i4, 3], 4)
)
```

Node 8 sits in six triangles and node 6 in three, yet node 6 scores higher ---
because node 8's triangles run through hubs, and a walk that wanders into a
hub is unlikely to come straight back. RWSE encodes "am I in a tight, *quiet*
neighbourhood", which is a genuinely different and often more useful signal
than the clustering coefficient. Just do not read it as one.

Laplacian PE says **where you are globally**; RWSE says **what your
neighbourhood looks like locally**. They are complementary, and GraphGPS uses
both.

::: {.callout-warning appearance="simple"}
## The sign problem

Eigenvectors are only defined up to sign: if $\mathbf{v}$ is an eigenvector so
is $-\mathbf{v}$. Nothing in the eigendecomposition picks one, so the same
graph can yield $2^k$ different Laplacian PEs, and a model that memorises one
will fail on another.

The standard fix is to randomly flip signs during training as an augmentation
(`sign_flip = TRUE` above), forcing sign-invariance. RWSE has no such ambiguity,
which is a real practical advantage.
:::

### 3. Attention bias --- the interesting one

The first two inject structure into $X$ and hope attention picks it up.
Graphormer [@ying2021transformers] modifies attention itself:

$$
S_{ij} = \frac{\mathbf{q}_i^\top \mathbf{k}_j}{\sqrt{d_h}} + b_{\phi(i,j)}
$$

where $\phi(i,j)$ is the shortest-path distance and $b$ is a **learnable scalar
per distance**. Add it before the softmax and every attention head sees the
graph's metric directly.

```{r}
spd_bias
```

Three things to appreciate about this:

-   It is $n^2$ terms of information --- far more than the $n \times k$ that
    positional encodings supply.
-   Attention stays **global**: unlike a GNN, node $i$ can attend to a node ten
    hops away if that is useful. The bias shapes the distribution rather than
    masking it. This sidesteps over-smoothing entirely, because there is no
    repeated local averaging.
-   Setting $b_{\phi} = 0$ for $\phi = 1$ and $-\infty$ otherwise recovers a
    GAT exactly. Graphormer is the soft version of a GNN's hard mask.

## Making it work

Here is the experiment. A planted-partition graph with three communities. We
reveal the true community label for 40% of nodes as an input feature and zero
out the rest; the model must infer the hidden labels by looking at the graph.

```{r}
#| label: task-setup
gs <- sbm_graph(n_per = 30, k = 3, p_in = 0.16, p_out = 0.01, seed = 11)
As <- as_adj_dense(gs)
n <- vcount(gs)
y <- as.integer(V(gs)$label)
K <- 3

# shortest-path distances, capped
maxd <- 5
D <- distances(gs)
D[is.infinite(D)] <- maxd
D <- pmin(D, maxd)
Didx <- as.integer(D) + 1L

set.seed(3)
revealed <- sample(n, round(0.4 * n))
hidden <- setdiff(seq_len(n), revealed)
train <- sample(hidden, round(length(hidden) / 2))
test <- setdiff(hidden, train)

Xlab <- matrix(0, n, K)
Xlab[cbind(revealed, y[revealed])] <- 1
X <- cbind(Xlab, lap_pe(As, 4), rwse(As, 1:4))
d <- ncol(X)

c(nodes = n, features = d, revealed = length(revealed),
  train = length(train), test = length(test))
```

The content-based part of attention is a random projection, and deliberately
uninformative — the label channels are zero for the majority of nodes, and
the structural channels do not encode community identity. The only route to
the answer is the graph.

```{r}
#| label: content-attn
set.seed(5)
Wq <- matrix(rnorm(d * d, sd = sqrt(1 / d)), d, d)
Wk <- matrix(rnorm(d * d, sd = sqrt(1 / d)), d, d)
S_content <- (X %*% Wq) %*% t(X %*% Wk) / sqrt(d)
```

### What the bias does to attention

Add $b_\phi = -\lambda \phi$ --- a simple linear decay in graph distance --- and
watch where attention goes. $\lambda = 0$ is the graph-blind transformer.

```{r}
#| label: lambda-sweep
same_comm <- outer(y, y, "==")
is_nbr <- As > 0

sweep_lambda <- function(lambda) {
  b <- -lambda * (0:maxd)
  Aw <- softmax_rows(S_content + matrix(b[Didx], n, n))
  c(`same community` = mean(rowSums(Aw * same_comm)),
    `direct neighbours` = mean(rowSums(Aw * is_nbr)),
    `entropy` = mean(-rowSums(Aw * log(Aw + 1e-12))))
}

round(t(sapply(c(0, 1, 2, 3), sweep_lambda)), 3)
```

Read the first column. With no bias, `r round(sweep_lambda(0)[1], 3)` of
attention mass lands on same-community nodes --- which is exactly chance, since
one third of the graph is your community. The transformer is blind. At
$\lambda = 2$ it is `r round(sweep_lambda(2)[1], 3)`. The row entropy falls from
`r round(sweep_lambda(0)[3], 2)` (against a maximum of
`r round(log(n), 2)` --- perfectly uniform) to
`r round(sweep_lambda(2)[3], 2)`.

One scalar per distance, added before the softmax, turned a set function into a
graph function.

```{r}
#| label: fig-attn
#| fig-cap: "Attention matrices, nodes ordered by community, self-attention removed and rows renormalised. Left: no bias — uniformly flat. Right: with a distance bias — the three communities appear as blocks on the diagonal."
#| fig-width: 8.5
#| fig-height: 4.3
plot_attention(
  list(softmax_rows(S_content),
       softmax_rows(S_content + matrix((-2 * (0:maxd))[Didx], n, n))),
  group = y,
  titles = c(expression(lambda == 0 ~ " (no graph)"),
             expression(lambda == 2 ~ " (distance bias)"))
)
```

The left panel is the whole problem in one image: every entry the same shade,
because a graph-blind transformer has no reason to prefer any node over any
other. The right panel is the fix.

### Does it help the task?

Attention mass is a means, not an end. Train a linear readout on top --- this
part has an exact gradient, so it is ordinary softmax regression:

```{r}
#| label: readout
fit_readout <- function(H, steps = 400, lr = 0.5) {
  Wo <- matrix(rnorm(ncol(H) * K, sd = 0.1), ncol(H), K)
  for (s in seq_len(steps)) {
    P <- softmax_rows(H %*% Wo)
    G <- P
    G[cbind(train, y[train])] <- G[cbind(train, y[train])] - 1
    G[-train, ] <- 0
    Wo <- Wo - lr * (t(H) %*% G) / length(train)
  }
  Wo
}

set.seed(9)
H_bias <- softmax_rows(S_content + matrix((-2 * (0:maxd))[Didx], n, n)) %*% X
H_none <- softmax_rows(S_content) %*% X

acc <- function(H, Wo, idx) mean(max.col(H %*% Wo)[idx] == y[idx])

Wb <- fit_readout(H_bias)
Wn <- fit_readout(H_none)

data.frame(
  model = c("attention + graph distance bias", "attention only (graph-blind)"),
  train = round(c(acc(H_bias, Wb, train), acc(H_none, Wn, train)), 3),
  test  = round(c(acc(H_bias, Wb, test),  acc(H_none, Wn, test)), 3)
)
```

`r round(acc(H_bias, Wb, test), 2)` against `r round(acc(H_none, Wn, test), 2)`.
Same architecture, same features, same readout, same number of parameters. The
only difference is one scalar per distance added to the attention logits.

That is the slide. Everything else in this book is the explanation of why that
number moves.

## Learning the bias

Graphormer does not hand-set $b$; it learns it. Ours has only
`r 6` parameters, so we can fit them by finite differences and skip autodiff
entirely --- perturb each parameter, see what happens to the loss.

```{r}
#| label: learn-bias
loss_fn <- function(par) {
  b <- par[1:(maxd + 1)]
  Wo <- matrix(par[-(1:(maxd + 1))], d, K)
  H <- softmax_rows(S_content + matrix(b[Didx], n, n)) %*% X
  P <- softmax_rows(H %*% Wo)
  -mean(log(pmax(P[cbind(train, y[train])], 1e-12)))
}

test_acc <- function(par) {
  b <- par[1:(maxd + 1)]
  Wo <- matrix(par[-(1:(maxd + 1))], d, K)
  H <- softmax_rows(S_content + matrix(b[Didx], n, n)) %*% X
  mean(max.col(H %*% Wo)[test] == y[test])
}

set.seed(9)
par <- c(rep(0, maxd + 1), rnorm(d * K, sd = 0.1))
eps <- 1e-4
trace <- numeric(200)

for (s in seq_len(200)) {
  gr <- numeric(length(par))
  for (j in seq_along(par)) {
    pp <- par; pp[j] <- pp[j] + eps
    pm <- par; pm[j] <- pm[j] - eps
    gr[j] <- (loss_fn(pp) - loss_fn(pm)) / (2 * eps)
  }
  par <- par - 1.0 * gr
  trace[s] <- loss_fn(par)
}

c(loss_start = round(trace[1], 3),
  loss_end = round(tail(trace, 1), 3),
  test_accuracy = round(test_acc(par), 3))
```

It learns --- the bias parameters started at zero, meaning graph-blind, and
gradient descent found its way to a working model.

Now look at *what* it learned:

```{r}
#| label: learned-bias
setNames(round(par[1:(maxd + 1)], 2), paste0("dist ", 0:maxd))
```

Not the smooth decay we hand-set. It is negative for distances 1–3 and then
strongly **positive** at distance 4.

This is not a bug, and it is worth sitting with. In a graph with three blocks,
a node four hops away is almost certainly in a *different* community --- and
"I am definitely not in that community" is, with three choices, nearly as
informative as "I am in this one". The readout has signed weights and can
happily invert the signal. The model found a strategy we did not anticipate,
and it is a correct one for this graph.

```{r}
Aw_learned <- softmax_rows(S_content + matrix(par[1:(maxd + 1)][Didx], n, n))
c(`same-community mass` = round(mean(rowSums(Aw_learned * same_comm)), 3),
  `chance` = round(1 / K, 3))
```

Confirmed: it attends *away* from its own community, well below chance.

The general lesson is the argument for learning the bias rather than fixing it:
the right shape depends on the graph and the task. On molecules, where
Graphormer was developed, chemistry is local and a decaying profile is the
sensible prior. On a block-structured social graph it is not. Leaving $b$
learnable lets the data decide, which is precisely why Graphormer parameterises
it that way.

::: {.callout-note appearance="simple"}
## What this demo is not

Finite differences cost two forward passes per parameter, which is fine for
`r 6 + 11 * 3` parameters and hopeless for a real model. The frozen random
$W_Q, W_K$ are doing no work here. A real graph transformer trains everything
jointly with backpropagation. The point of this version is that you can read
every line of it.
:::

## Doing it for real, in R

The R route to actual graph transformers is the `torch` package --- a direct
binding to libtorch, no Python involved. The pieces we built map over
one-for-one:

```{r}
#| eval: false
library(torch)

graph_attention_bias <- nn_module(
  "graph_attention_bias",
  initialize = function(d_model, n_heads, max_dist = 8) {
    self$n_heads <- n_heads
    self$max_dist <- max_dist
    self$attn <- nn_multihead_attention(d_model, n_heads, batch_first = TRUE)
    # one learnable scalar per (head, distance) -- Graphormer's spatial encoding
    self$spatial <- nn_embedding(max_dist + 1, n_heads)
  },
  forward = function(x, dist) {
    # x: (batch, n, d_model);  dist: (batch, n, n) integer hop distances
    d <- torch_clamp(dist, max = self$max_dist)
    bias <- self$spatial(d)$permute(c(1, 4, 2, 3))   # (batch, heads, n, n)
    b <- bias$reshape(c(-1, dist$size(2), dist$size(3)))
    self$attn(x, x, x, attn_mask = b)[[1]]
  }
)
```

Precompute `dist` once per graph with `igraph::distances()` and cache it; it is
static, and recomputing all-pairs shortest paths every batch will dominate your
training time.

For the input-embedding route (option 1), no custom module is needed at all:

```{r}
#| eval: false
Z <- netmf(A, dim = 128)                     # embed the entity graph offline
emb <- nn_embedding(nrow(Z), 128)
emb$weight <- nn_parameter(torch_tensor(Z))  # initialise, then fine-tune
```

## The catch

Graph transformers have a hard limit that GNNs do not: the attention matrix is
$n \times n$. A GNN costs $O(|E|)$ per layer; a graph transformer costs
$O(n^2)$. For molecules (tens of atoms) that is nothing. For a social network
with a million nodes it is a trillion entries.

Hence the dominant architecture is hybrid --- **GraphGPS**
[@rampasek2022recipe] runs a message-passing layer and an attention layer in
parallel at each block, summing the outputs. The MPNN handles local structure
cheaply over edges; attention handles long-range interactions; positional
encodings glue them together. If you build a graph transformer today, start
there.

## Summary

| Approach | Use when | Cost |
|---|---|---|
| Pretrained embeddings as input vectors | Tokens are entities in a side graph | Free |
| Laplacian PE | Global position matters; watch the sign ambiguity | $O(n k)$ |
| RWSE | Local structure matters; no sign issues | $O(nK)$ dense, cheap sparse |
| Attention bias | The graph *is* the input, $n$ is small | $O(n^2)$ |
| Hybrid (GraphGPS) | Large graphs, long-range interactions | $O(|E| + n^2)$ per block |
