# Evaluating and choosing {#sec-eval}

```{r}
#| label: setup
source("R/graphemb.R")
set.seed(1)
```

An embedding has no intrinsic quality. It is only good *for something*, and the
main way people fool themselves is by never checking which something.

## Link prediction

The standard intrinsic benchmark, because it needs no labels. Hide a fraction
of the edges, embed what is left, and check whether the hidden edges score
higher than random non-edges.

The critical detail --- and the most common bug in hand-rolled evaluation ---
is that **the held-out edges must be removed before embedding**. Embed the full
graph and then "predict" edges the embedder already saw and you will report an
AUC near 1.0 and be very pleased with yourself.

```{r}
edge_split
```

```{r}
link_pred_auc
```

```{r}
auc
```

We use a random geometric graph: points in the unit square joined when close.
It has a genuine latent structure to recover, unlike an SBM where edges are
conditionally independent given community and the achievable AUC is capped well
below 1.

```{r}
#| label: linkpred
gg <- geo_graph(n = 300, radius = 0.115, seed = 7)
sp <- edge_split(gg, frac = 0.15, seed = 3)
A_tr <- as_adj_dense(sp$g_train)

c(nodes = vcount(gg), edges = ecount(gg), held_out = nrow(sp$pos))
```

Now compare methods --- including, importantly, two baselines that involve no
embedding at all.

```{r}
#| label: linkpred-compare
common_neighbours <- function(A, e) rowSums(A[e[, 1], ] * A[e[, 2], ])

adamic_adar <- function(A, e) {
  w <- 1 / log(pmax(rowSums(A), 2))
  rowSums(A[e[, 1], ] * A[e[, 2], ] * rep(w, each = nrow(e)))
}

heuristic_auc <- function(f) {
  s <- c(f(A_tr, sp$pos), f(A_tr, sp$neg))
  auc(s, c(rep(1, nrow(sp$pos)), rep(0, nrow(sp$neg))))
}

results <- data.frame(
  method = c("Common neighbours", "Adamic–Adar", "Spectral (d=16)",
             "NetMF (d=32)", "DeepWalk (d=32)"),
  auc = round(c(
    heuristic_auc(common_neighbours),
    heuristic_auc(adamic_adar),
    link_pred_auc(spectral_embedding(A_tr, d = 16), sp$pos, sp$neg),
    link_pred_auc(netmf(A_tr, dim = 32), sp$pos, sp$neg),
    link_pred_auc(deepwalk(A_tr, dim = 32, num_walks = 5, walk_length = 30,
                           epochs = 5, seed = 2), sp$pos, sp$neg)
  ), 3)
)
results[order(-results$auc), ]
```

::: {.callout-important appearance="simple"}
## Always run the heuristic baseline

Common neighbours is one line of code, has no hyperparameters, and takes
microseconds. On this graph it is competitive with every embedding method.

That is not an artefact of the toy size --- it is a real and frequently
reproduced result on link prediction benchmarks. If your embedding pipeline
cannot beat "count the mutual friends", the pipeline is not earning its
complexity. Run the baseline first, every time.
:::

The embeddings do have one advantage the heuristics do not: they produce a
fixed-width vector per node that you can feed to anything. Adamic–Adar gives
you a score for a *pair*, and nothing else.

## Node classification

If you have labels, the question is whether the embedding makes them
recoverable. Use a deliberately weak classifier --- the point is to measure the
representation, not to win a leaderboard.

```{r}
knn_accuracy
```

```{r}
#| label: nodeclass
gs <- sbm_graph(n_per = 40, k = 3, p_in = 0.10, p_out = 0.015, seed = 5)
As <- as_adj_dense(gs)
ls <- V(gs)$label

data.frame(
  method = c("Spectral (d=8)", "NetMF (d=32)", "DeepWalk (d=32)",
             "Random GCN (2 layers)"),
  knn_accuracy = round(c(
    knn_accuracy(spectral_embedding(As, d = 8), ls, k = 5),
    knn_accuracy(netmf(As, dim = 32), ls, k = 5),
    knn_accuracy(deepwalk(As, dim = 32, epochs = 5, seed = 1), ls, k = 5),
    knn_accuracy(gcn_forward(As, diag(nrow(As)),
                             gcn_random_weights(c(nrow(As), 32, 8), seed = 2)),
                 ls, k = 5)
  ), 3)
)
```

This graph is deliberately harder than the one in @sec-spectral --- sparser
communities, more cross-edges --- so the methods separate instead of all
saturating at 1.0.

## The evaluation traps

Five ways to get a number that does not mean what you think.

**1. Leakage through the graph.** Covered above, and worth repeating because it
is the one that silently produces publishable-looking results.

**2. Degree as a confounder.** Both link prediction and classification are
easier for high-degree nodes. A method that only learned degree can look
respectable. Check by stratifying:

```{r}
#| label: degree-strat
Z <- netmf(A_tr, dim = 32)
deg <- rowSums(A_tr)
pos_deg <- rowMeans(cbind(deg[sp$pos[, 1]], deg[sp$pos[, 2]]))
grp <- cut(pos_deg, quantile(pos_deg, c(0, 1/3, 2/3, 1)),
           labels = c("low", "mid", "high"), include.lowest = TRUE)

sapply(levels(grp), function(gl) {
  keep <- grp == gl
  round(link_pred_auc(Z, sp$pos[keep, , drop = FALSE], sp$neg), 3)
})
```

If performance were driven entirely by degree, the low-degree stratum would
collapse. Here it holds up, which is the reassuring outcome.

**3. Negative sampling that is too easy.** Uniformly sampled non-edges are
usually far apart in the graph and trivial to reject. Harder and more
informative: sample negatives at distance exactly 2.

```{r}
#| label: hard-negatives
D_tr <- distances(sp$g_train)
cand <- which(D_tr == 2, arr.ind = TRUE)
cand <- cand[cand[, 1] < cand[, 2], ]
set.seed(4)
hard_neg <- cand[sample.int(nrow(cand), nrow(sp$pos)), ]

c(easy_negatives = round(link_pred_auc(Z, sp$pos, sp$neg), 3),
  hard_negatives = round(link_pred_auc(Z, sp$pos, hard_neg), 3))
```

The AUC drops sharply. Both numbers are "correct"; they answer different
questions, and only the second resembles a deployed ranking problem, where
you are choosing among plausible candidates rather than against random noise.

**4. Comparing at unequal dimension.** A $d = 128$ embedding against a $d = 16$
one is not a comparison of methods. Match $d$, or sweep it.

**5. Single runs of stochastic methods.** DeepWalk depends on walk sampling,
SGD ordering, and initialisation. Report a spread:

```{r}
#| label: variance
set.seed(1)
accs <- sapply(1:5, function(s) {
  knn_accuracy(deepwalk(As, dim = 32, epochs = 5, seed = s), ls, k = 5)
})
c(round(accs, 3), mean = round(mean(accs), 3), sd = round(sd(accs), 3))
```

If two methods differ by less than this spread, they do not differ.

## Choosing a method

```{r}
#| label: fig-decision
#| echo: false
#| fig-cap: "A rough decision procedure. Start at the top."
#| fig-width: 7.5
#| fig-height: 4.6
par(mar = c(0, 0, 0, 0))
plot.new(); plot.window(c(0, 10), c(0, 10))
bx <- function(x, y, w, h, txt, fill) {
  rect(x - w/2, y - h/2, x + w/2, y + h/2, col = fill, border = "grey40")
  text(x, y, txt, cex = 0.78)
}
ar <- function(x0, y0, x1, y1, lab = NULL) {
  arrows(x0, y0, x1, y1, length = 0.08, col = "grey40")
  if (!is.null(lab)) text((x0 + x1)/2, (y0 + y1)/2, lab, cex = 0.68,
                          col = "grey25", pos = 4, offset = 0.2)
}
bx(5, 9.3, 5.2, 1.0, "Do nodes have features\nthat matter?", "#EAF0F7")
bx(2, 6.9, 3.4, 1.0, "Does the graph\nchange?", "#EAF0F7")
bx(8, 6.9, 3.4, 1.0, "GNN\n(GCN / GraphSAGE)", "#DDEBD8")
ar(3.6, 9.0, 2.6, 7.5, "no"); ar(6.4, 9.0, 7.4, 7.5, "yes")
bx(2, 4.4, 3.4, 1.0, "GNN or\nre-embed on a schedule", "#DDEBD8")
bx(6, 4.4, 3.6, 1.0, "n < ~20k ?", "#EAF0F7")
ar(1.4, 6.3, 1.4, 5.0, "yes"); ar(2.9, 6.4, 5.2, 4.9, "no")
bx(4.2, 1.9, 3.2, 1.0, "NetMF\n(closed form)", "#DDEBD8")
bx(8.2, 1.9, 3.2, 1.0, "DeepWalk / node2vec\n(sampled)", "#DDEBD8")
ar(5.3, 3.9, 4.4, 2.5, "yes"); ar(6.9, 3.9, 7.9, 2.5, "no")
```

Written out, with the caveats that matter:

| Situation | Use | Why |
|---|---|---|
| First thing you try, any graph | **Spectral** | Deterministic, no tuning, tells you the number of communities |
| Fixed graph, no features, $n \lesssim$ 20k | **NetMF** | Closed form, matches DeepWalk, milliseconds |
| Fixed graph, no features, large $n$ | **DeepWalk / node2vec** | Sampling is what scales |
| Need control over proximity radius | **node2vec** | $p, q$ tune reach --- but see @sec-walks on what they do *not* do |
| Node features matter | **GCN / GraphSAGE** | Only family that uses them |
| Graph grows, new nodes need embedding | **GraphSAGE** | Inductive; shallow methods cannot do this at all |
| Graph *is* the model input, $n$ small | **Graph transformer** | Global receptive field, no over-smoothing |
| Large graph, long-range interactions | **GraphGPS hybrid** | $O(\|E\|)$ locally, attention globally |

And the honest addendum: **try the heuristic first**. Common neighbours,
Adamic–Adar, personalised PageRank, and plain degree/triangle features solve a
surprising number of graph problems, and they give you the number your
embedding has to beat.

## Practical notes for R

**Sparse matrices.** Use **Matrix**. Everything in this book used dense
matrices for readability, which is fine to about 10,000 nodes and catastrophic
beyond.

**Eigendecomposition.** **RSpectra**'s `eigs_sym()` for the few smallest
eigenvalues; **irlba** for truncated SVD. Never `eigen()` or `svd()` on
anything large --- they compute the full decomposition.

**Graph manipulation.** **igraph** for essentially everything.

**Existing implementations.** R has no equivalent of PyTorch Geometric.
Realistically: implement the shallow methods yourself (they are short, as this
book demonstrates), use **word2vec** or **rsparse** for the SGNS step if you
want a compiled trainer, and use **torch** or **reticulate** onto PyG/DGL for
GNNs and graph transformers.

**Visualisation.** **uwot** (UMAP) or **Rtsne** to get from $d = 128$ down to 2
for plotting. Do not use the 2D projection as your embedding --- project for
your eyes, model on the full-dimensional vectors.

```{r}
#| eval: false
library(uwot)
Z <- netmf(A, dim = 128)
emb2d <- umap(Z, n_neighbors = 15, min_dist = 0.1)
plot(emb2d, col = factor(labels), pch = 19, cex = 0.5)
```

## What to remember

1.  Hold out edges **before** embedding.
2.  Run the heuristic baseline.
3.  Match dimensions when comparing.
4.  Report a spread over seeds for anything stochastic.
5.  Stratify by degree, or you may just be measuring degree.
6.  Choose the evaluation that matches your actual task --- and if you have a
    downstream task, evaluate on *it*, not on link prediction.
