---
title: "Graph Embeddings in R"
subtitle: "From Laplacians to transformers, in 7 minutes"
author: "Adam Black"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    fig-align: center
    width: 1280
    height: 760
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
source(if (file.exists("R/graphemb.R")) "R/graphemb.R" else "../R/graphemb.R")
set.seed(1)

g <- karate_graph()
A <- as_adj_dense(g)
lab <- V(g)$label
blue <- "#4C72B0"; orange <- "#DD8452"; green <- "#55A868"
fac_col <- ifelse(lab == "Mr Hi", blue, orange)
```

## The problem {.smaller}

::: {.columns}
::: {.column width="52%"}
```{r}
#| fig-width: 5.4
#| fig-height: 5
set.seed(4)
par(mar = c(0, 0, 0, 0))
plot(g, layout = layout_with_fr(g), vertex.color = fac_col,
     vertex.label.color = "white", vertex.size = 16,
     vertex.label.cex = 0.75, edge.color = "grey75")
```
:::
::: {.column width="48%"}
A graph: 34 people, 78 friendships, and a club that split in two.

Every model you want to fit needs **a table**. You have **an adjacency matrix**.

It is the wrong shape in three ways:

- 34 columns wide --- changes when someone joins
- `r round(100 * mean(A == 0))`% zeros
- **no geometry** --- 3 hops away in the same clique looks as unrelated as a different continent

::: {.fragment}
An adjacency matrix records *edges*. You want *similarity* --- and in a graph that comes from **paths**.
:::
:::
:::

## The idea {.smaller}

::: {.columns}
::: {.column width="46%"}
A map $f: V \to \mathbb{R}^d$ from nodes to short vectors, so that

> **close in the vector space**
> = **structurally close in the graph**

$d$ is 16–256. Fixed, dense, independent of $n$.

::: {.fragment}
Colour = the faction they actually joined.
The algorithm **never saw the labels**.

Axis 1 alone gets **33 of 34** right.

Node 9 --- who joined the officers despite being better connected to the instructor --- lands in the middle. The honest answer.
:::
:::
::: {.column width="54%"}
```{r}
#| fig-width: 6
#| fig-height: 4.8
Z <- spectral_embedding(A, d = 2)
par(mar = c(4, 4, 1, 1))
plot(Z[, 1], Z[, 2], type = "n", xlab = "dimension 1", ylab = "dimension 2")
abline(v = 0, col = "grey85", lty = 2)
text(Z[, 1], Z[, 2], labels = seq_len(nrow(Z)), cex = 0.9, col = fac_col)
```
:::
:::

## 1. Spectral --- four lines {.smaller}

Want connected nodes close? Penalise every edge by the gap it stretches:

$$
\tfrac{1}{2}\sum_{i,j} A_{ij}(f_i - f_j)^2 \;=\; \mathbf{f}^\top \underbrace{(D - A)}_{\text{Laplacian } L} \mathbf{f}
$$

::: {.fragment}
Minimise, constrain so it can't collapse → **eigenvectors of $L$ with the smallest eigenvalues.**
The smoothest non-constant functions on the graph.

```r
L  <- diag(rowSums(A)) - A
ev <- eigen(L, symmetric = TRUE)
Z  <- ev$vectors[, (n-1):(n-d)]     # drop the trivial constant vector
```
:::

::: {.fragment}
**Free bonus:** a gap in the eigenvalues tells you how many communities there are.

```{r}
#| echo: false
g3 <- sbm_graph(n_per = 40, k = 3, p_in = 0.14, p_out = 0.01, seed = 42)
round(sort(eigen(graph_laplacian(as_adj_dense(g3), "sym"), symmetric = TRUE)$values)[1:6], 3)
```

Three near zero, then a jump. Three communities.
:::

## 2. Random walks --- DeepWalk {.smaller}

::: {.columns}
::: {.column width="55%"}
2014: word2vec learns word vectors from *"words that appear near each other are related"*.

**A random walk on a graph is a sentence.**

```r
walks <- random_walks(A, num_walks = 10,
                      walk_length = 40)
pairs <- walks_to_pairs(walks, window = 5)
fit   <- skipgram_sgns(pairs, n, dim = 32)
```

Skip-gram with negative sampling:

$$
-\log \sigma(\mathbf{w}_u^\top \mathbf{c}_v) - \sum_{k=1}^{K} \log \sigma(-\mathbf{w}_u^\top \mathbf{c}_{v'_k})
$$

Real pairs score high, random pairs low. **~60 lines of base R, no autodiff.**
:::
::: {.column width="45%"}
```{r}
#| fig-width: 5
#| fig-height: 4.3
Z_dw <- deepwalk(A, dim = 32, epochs = 5, seed = 1)
px <- prcomp(Z_dw)$x[, 1:2]
par(mar = c(4, 4, 2, 1))
plot(px, type = "n", xlab = "PC1", ylab = "PC2",
     main = sprintf("5-NN accuracy: %.2f", knn_accuracy(Z_dw, lab, 5)))
text(px[, 1], px[, 2], labels = seq_len(nrow(A)), cex = 0.85, col = fac_col)
```

Nothing is $O(n^2)$ → this is the family that scaled to **billions of edges**.
:::
:::

## 3. They're all the same algorithm {.smaller}

::: {.callout-important appearance="minimal"}
**DeepWalk is not really a neural method.** Skip-gram provably converges to a factorisation of a PMI matrix (Levy & Goldberg 2014), and for random walks that matrix is closed-form (Qiu et al. 2018):

$$
M = \frac{\mathrm{vol}(G)}{bT}\Big(\sum_{r=1}^{T} P^r\Big) D^{-1}, \qquad P = D^{-1}A
$$
:::

::: {.columns}
::: {.column width="50%"}
```{r}
#| echo: false
tab <- data.frame(
  Method = c("Spectral", "DeepWalk / NetMF", "node2vec", "LINE"),
  `Factorises` = c("$I - D^{-1/2}AD^{-1/2}$",
                   "$\\log\\max(M, 1)$",
                   "$M$ with 2nd-order $P$",
                   "$M$ with $T = 1$"),
  check.names = FALSE
)
knitr::kable(tab)
```

Every method: **a low-rank factorisation of a matrix built from $A$.**
They differ in *which matrix*.
:::
::: {.column width="50%"}
```{r}
#| echo: false
t_dw <- system.time(Zd <- deepwalk(A, dim = 32, epochs = 5, seed = 1))[["elapsed"]]
t_nm <- system.time(Zn <- netmf(A, dim = 32))[["elapsed"]]
knitr::kable(data.frame(
  ` ` = c("DeepWalk (sampled)", "NetMF (SVD)"),
  seconds = signif(c(t_dw, t_nm), 3),
  `5-NN acc` = c(knn_accuracy(Zd, lab, 5), knn_accuracy(Zn, lab, 5)),
  check.names = FALSE
))
```

::: {.fragment}
Same answer. **~1000× faster.**

Before you reach for a library: **try an SVD.**
:::
:::
:::

## 4. GNNs --- when $A$ isn't enough {.smaller}

Everything above is a function of $A$ **alone**. So it cannot use node features, and cannot embed a node that wasn't there.

::: {.fragment}
GNNs replace the lookup table with a **function of the neighbourhood**:

$$
H^{(l+1)} = \sigma\big(\hat{A}\,H^{(l)}\,W^{(l)}\big), \qquad \hat{A} = \tilde D^{-1/2}(A+I)\tilde D^{-1/2}
$$

Weights are **shared**, so a new node is just a forward pass.
:::

::: {.fragment}
Two things to know:

- A GCN layer is a **learned smoother**. Stack 20 and every node converges to the same vector → GNNs are 2–3 layers deep.
- **GAT = self-attention masked by the adjacency matrix.**
:::

::: {.fragment .highlight}
So what happens if you take the mask *off*?
:::

## Into a transformer: three doors {.smaller}

Self-attention is **permutation-invariant** --- permute the inputs, the outputs permute too. That's why transformers need positional encodings. Text has an order. **A graph doesn't.**

::: {.columns}
::: {.column width="33%"}
### 1. Input vectors
Tokens that are entities in a side graph --- clinical codes, products, users.

Embed the graph offline, initialise the embedding table.

**Zero architecture change.**
:::
::: {.column width="33%"}
### 2. Positional encodings
Give each node its Laplacian eigenvectors --- *literally slide 4*.

Sine waves are the eigenvectors of the Laplacian of a line graph. The analogy is exact.
:::
::: {.column width="33%"}
### 3. Attention bias
Put the graph **inside** the attention computation.

$$
S_{ij} = \frac{\mathbf{q}_i^\top\mathbf{k}_j}{\sqrt{d_h}} + b_{\phi(i,j)}
$$

$\phi$ = shortest-path distance.
:::
:::

## The result {.smaller}

```{r}
#| label: demo
#| include: false
gs <- sbm_graph(n_per = 30, k = 3, p_in = 0.16, p_out = 0.01, seed = 11)
As <- as_adj_dense(gs); n <- vcount(gs); y <- as.integer(V(gs)$label); K <- 3
maxd <- 5
D <- distances(gs); D[is.infinite(D)] <- maxd; D <- pmin(D, maxd)
Didx <- as.integer(D) + 1L
set.seed(3)
revealed <- sample(n, round(0.4 * n)); hidden <- setdiff(seq_len(n), revealed)
train <- sample(hidden, round(length(hidden) / 2)); test <- setdiff(hidden, train)
Xlab <- matrix(0, n, K); Xlab[cbind(revealed, y[revealed])] <- 1
X <- cbind(Xlab, lap_pe(As, 4), rwse(As, 1:4)); d <- ncol(X)
set.seed(5)
Wq <- matrix(rnorm(d * d, sd = sqrt(1 / d)), d, d)
Wk <- matrix(rnorm(d * d, sd = sqrt(1 / d)), d, d)
Sc <- (X %*% Wq) %*% t(X %*% Wk) / sqrt(d)
same <- outer(y, y, "==")

fit_readout <- function(H, steps = 400, lr = 0.5) {
  Wo <- matrix(rnorm(ncol(H) * K, sd = 0.1), ncol(H), K)
  for (s in seq_len(steps)) {
    P <- softmax_rows(H %*% Wo); G <- P
    G[cbind(train, y[train])] <- G[cbind(train, y[train])] - 1
    G[-train, ] <- 0
    Wo <- Wo - lr * (t(H) %*% G) / length(train)
  }
  Wo
}
set.seed(9)
Aw0 <- softmax_rows(Sc)
Aw2 <- softmax_rows(Sc + matrix((-2 * (0:maxd))[Didx], n, n))
H0 <- Aw0 %*% X; H2 <- Aw2 %*% X
W0 <- fit_readout(H0); W2 <- fit_readout(H2)
acc <- function(H, W) mean(max.col(H %*% W)[test] == y[test])
a0 <- acc(H0, W0); a2 <- acc(H2, W2)
m0 <- mean(rowSums(Aw0 * same)); m2 <- mean(rowSums(Aw2 * same))
```

::: {.columns}
::: {.column width="58%"}
```{r}
#| fig-width: 7
#| fig-height: 3.9
plot_attention(list(Aw0, Aw2), group = y,
               titles = c("no graph", "+ distance bias"))
```

90 nodes, 3 communities. Reveal 40% of labels, infer the rest.
The content part of attention is deliberately uninformative --- **the only route is the graph.**
:::
::: {.column width="42%"}
Attention mass on your **own community**:

| | |
|---|---|
| no graph | `r sprintf("%.0f%%", 100 * m0)` *(= chance)* |
| + bias | **`r sprintf("%.0f%%", 100 * m2)`** |

::: {.fragment}
Downstream **test accuracy**:

| | |
|---|---|
| no graph | `r sprintf("%.0f%%", 100 * a0)` |
| + bias | **`r sprintf("%.0f%%", 100 * a2)`** |

Same architecture. Same features. Same parameter count.

**One scalar per distance, added before the softmax.**
:::
:::
:::

## Takeaways

::: {.incremental}
1. **Graph embeddings turn structure into geometry.** Small vectors, ordinary models downstream --- and the algorithms are *short*. All four here are base R.

2. **They're all the same algorithm:** a low-rank factorisation of a matrix built from $A$. Try an SVD before a library --- and always run *"count the mutual friends"* as a baseline. It wins more often than anyone admits.

3. **If a transformer is near your graph, the structure has to get in somewhere.** Input vectors, positional encodings, or attention bias. Pick at least one.
:::

::: {.fragment}
::: {.callout-note appearance="minimal"}
Book, code, and every number on these slides:
**github.com/ablack3033/graph-embeddings-talk**
:::
:::
