---
title: "Graph Embeddings 101"
subtitle: "or: How I Learned to Stop Worrying and Love Claude Code"
author: "Adam Black"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    fig-align: center
    width: 1280
    height: 760
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
source(if (file.exists("R/graphemb.R")) "R/graphemb.R" else "../R/graphemb.R")
set.seed(1)

g <- karate_graph()
A <- as_adj_dense(g)
lab <- V(g)$label
blue <- "#4C72B0"; orange <- "#DD8452"; green <- "#55A868"
fac_col <- ifelse(lab == "Mr Hi", blue, orange)
```

# Part 0 {.center}

### Two minutes of unrelated grievances

## The title is a reference {.smaller}

::: {.columns}
::: {.column width="55%"}
*Dr. Strangelove, or: How I Learned to Stop Worrying and Love the Bomb* (1964).

Kubrick set out to make a serious thriller about nuclear annihilation, read the
actual strategy documents, and discovered they were **already comedies**. So he
gave up and filmed that instead.

::: {.fragment}
I have had a similar experience reading AI discourse.
:::
:::
::: {.column width="45%"}
::: {.callout-note appearance="minimal"}
### The plot, briefly
The Soviets build a Doomsday Machine that destroys the world automatically and
cannot be switched off.

It has one flaw.

::: {.fragment}
**They forgot to announce it.**

A deterrent nobody knows about deters nobody.
:::
:::
:::
:::

::: {.fragment .highlight}
Which is why every AI lab has a blog.
:::

## The bomb comparison, briefly {.smaller}

Everyone reaches for it, so let me get it out of the way and be precise about
which parts hold:

::: {.columns}
::: {.column width="50%"}
### Holds
- Physicists knew roughly what it would do and built it anyway
- "If we don't, they will" is doing enormous load-bearing work in both eras
- The people closest to it are the most nervous
:::
::: {.column width="50%"}
### Does not hold
- The Manhattan Project was about **200 people** and one government
- It did not ship as a **\$20/month subscription**
- Nobody had to explain Trinity to their parents at Thanksgiving
:::
:::

::: {.fragment}
Oppenheimer got *"Now I am become Death, destroyer of worlds."*
:::

::: {.fragment .highlight}
We got "You're absolutely right!"
:::

::: {.fragment}
::: {.callout-important appearance="minimal"}
In the film, *"I learned to stop worrying"* is sarcasm. I am choosing to read my
version straight. Ask me again in five years.
:::
:::

## Hi, I'm Adam {.smaller}

::: {.incremental}
- I am new to Boston. I moved here from the Netherlands.
- Last week I went to the DMV to get my license.
- Before that I got **two parking tickets** — for parking outside my own house.
  While moving into it. With the boxes still in the car.
:::

::: {.fragment}
So I asked the guy: *what's more expensive — the tickets, or just paying for parking?*
:::

::: {.fragment .highlight}
"Paying for parking. Welcome to Boston."
:::

## The bike lanes {.smaller}

I want to be fair. **For an American city, Boston is genuinely not bad.**

That is the nicest available sentence and I have thought about it a lot.

::: {.columns}
::: {.column width="50%"}
### Rotterdam
```{=html}
<svg viewBox="0 0 420 150" width="100%" role="img" aria-label="Cross-section of a Dutch protected bike lane, separated from traffic by a raised kerb">
  <rect x="0" y="60" width="150" height="60" fill="#5b6670"/>
  <text x="75" y="96" font-size="15" fill="#fff" text-anchor="middle">cars</text>
  <rect x="150" y="52" width="34" height="68" fill="#8fb08a"/>
  <text x="167" y="142" font-size="12" fill="#3b6b3b" text-anchor="middle">kerb</text>
  <rect x="184" y="52" width="130" height="68" fill="#b23b3b"/>
  <text x="249" y="92" font-size="15" fill="#fff" text-anchor="middle">bikes</text>
  <rect x="314" y="44" width="106" height="76" fill="#c9c4bb"/>
  <text x="367" y="88" font-size="14" fill="#333" text-anchor="middle">people</text>
</svg>
```
Three separate places, at three different heights, in three different colours.
No negotiation required.
:::
::: {.column width="50%"}
### Boston
```{=html}
<svg viewBox="0 0 420 150" width="100%" role="img" aria-label="Cross-section of an American painted bike lane: a stripe of paint between moving traffic and parked cars">
  <rect x="0" y="60" width="230" height="60" fill="#5b6670"/>
  <text x="115" y="96" font-size="15" fill="#fff" text-anchor="middle">cars</text>
  <rect x="230" y="60" width="60" height="60" fill="#68737d"/>
  <line x1="232" y1="60" x2="232" y2="120" stroke="#f2f2f2" stroke-width="4" stroke-dasharray="9 7"/>
  <text x="260" y="95" font-size="12" fill="#fff" text-anchor="middle">bikes*</text>
  <rect x="290" y="60" width="60" height="60" fill="#5b6670"/>
  <text x="320" y="96" font-size="11" fill="#fff" text-anchor="middle">parked</text>
  <rect x="350" y="44" width="70" height="76" fill="#c9c4bb"/>
  <text x="385" y="88" font-size="13" fill="#333" text-anchor="middle">people</text>
</svg>
```
One flat surface. The bike lane is a **paint colour**, and it is stored between
moving traffic and the doors of parked cars.
:::
:::

::: {.fragment}
\* Also where the delivery van goes. And the police car. And the cones.
:::

::: {.notes}
Swap both SVGs for real photos if you can get them — a Rotterdam protected lane
next to a Boston painted gutter lane gets a much bigger laugh than a diagram.
See slides/README-images.md for what to shoot and where to drop the files.
:::

## It's not the infrastructure {.smaller}

The lanes are the easy half. The hard half is that in the Netherlands
**everyone knows the protocol.**

A pedestrian hears a bike, so they keep walking. The bike routes around them.
Nobody makes eye contact. It just flows. It is a distributed system with no
coordinator and it is beautiful.

::: {.fragment}
Here, someone steps into the bike lane, sees me coming, and —
:::

::: {.fragment .highlight}
freezes.
:::

::: {.fragment}
In the bike lane. Full eye contact. Waiting for **me** to decide what happens
next.
:::

::: {.fragment}
And I have come to think this is the most American thing I have experienced:
a total stranger, in a moment of real physical danger, concluding that the
correct move is **synchronous negotiation**.
:::

## Anyway {.center}

::: {.fragment .highlight}
We're here to talk about graph embeddings.
:::

::: {.fragment}
And, apparently, Claude.
:::

# Part 1 {.center}

### Graph embeddings, from nothing

## A confession {.smaller}

::: {.fragment}
Before I started writing this talk, I knew **very little** about graph
embeddings.
:::

::: {.fragment}
I did not pick the topic because I'm an expert. I picked it because I wanted to
*become* one, and agreeing to give a talk is the only reliable forcing function
I have ever found.
:::

::: {.fragment}
So this is a talk about something I learned recently — which is the best kind,
because I still remember which parts were confusing.
:::

::: {.fragment .highlight}
I learned it from Claude. It is a genuinely excellent teacher.
:::

## How I actually study now {.smaller}

::: {.columns}
::: {.column width="55%"}
1. Open-ended conversation about a thing I don't understand
2. *"Turn that into a Quarto book — one chapter per idea, **runnable code in
   every chapter**"*
3. *"Now write me a lecture script from chapter 3"*
4. Pipe the script through the ElevenLabs API
5. Listen to it on the bike. Well — on the **T**. I'm not doing the bike lane
   thing again.
:::
::: {.column width="45%"}
::: {.callout-tip appearance="minimal"}
### The step that matters
**Runnable code in every chapter.**

A book that only produces prose will happily explain something false to you for
forty pages, fluently.

Code that has to run is a lie detector.
:::
:::
:::

::: {.fragment}
Here is a clip of a synthetic voice — one that sounds *suspiciously* like a
certain Oscar winner — patiently explaining the **Riesz representation theorem**
and Hilbert spaces to me at 1.4× speed.
:::

::: {.fragment .highlight}
Donald knows what I'm talking about.
:::

::: {.notes}
Clip text is `slides/riesz-clip.txt` (6–7 min in full; the long-form version
with references is `slides/riesz-lecture.md`).

On stage play only the **first paragraph, ~20 seconds** — cut on "the difference
between a thing, and a measurement of a thing." That is the whole bit: it sounds
like a real lecture because it is one.

Don't oversell it. The laugh is entirely in the contrast between a serene,
expensive-sounding narrator and the words "Riesz representation theorem", so any
calm stock voice works. Use one you have rights to; don't clone a named person
for a talk that gets recorded and posted.
:::

## Start from the top: what is a graph? {.smaller}

::: {.columns}
::: {.column width="50%"}
A set of **nodes** $V$ and a set of **edges** $E \subseteq V \times V$.

That's it. That is the whole object.

::: {.fragment}
Natural question: **is it a metric space?**

Probably yes, right? Because we can define a distance.
:::

::: {.fragment}
A metric is any $d: V \times V \to \mathbb{R}$ with

- $d(x,y) \ge 0$, and $= 0$ only when $x = y$
- $d(x,y) = d(y,x)$
- $d(x,z) \le d(x,y) + d(y,z)$
:::
:::
::: {.column width="50%"}
```{r}
#| fig-width: 5.2
#| fig-height: 4.4
set.seed(4)
par(mar = c(0, 0, 0, 0))
plot(g, layout = layout_with_fr(g), vertex.color = fac_col,
     vertex.label.color = "white", vertex.size = 16,
     vertex.label.cex = 0.75, edge.color = "grey75")
```

34 people, 78 friendships, one club that split in two.
:::
:::

## Which distance, though? {.smaller}

Plenty of functions satisfy those three axioms, and they disagree with each other:

| | what it measures | trouble |
|---|---|---|
| **shortest path** | fewest hops | integer-valued, and everything is 2–6 |
| **commute time** | how long a random walk takes to get there and back | uses *all* paths, not the best one |
| **resistance** | the graph as an electrical circuit | degrades on large sparse graphs |
| **mutual friends** | overlap of two neighbourhoods | not a metric; embarrassingly effective |

::: {.fragment}
So we do have a metric space. The problem was never that we lacked a distance —
:::

::: {.fragment .highlight}
it's that a distance isn't a *coordinate*.
:::

::: {.fragment}
`lm()` cannot take a distance. Neither can a neural network. They want a **row
of numbers**.
:::

## What is an embedding? {.smaller}

::: {.columns}
::: {.column width="46%"}
Precisely: a map
$$f: V \to \mathbb{R}^d$$
that represents each point as a **vector**, such that

> **close in the vector space**
> = **structurally close in the graph**

$d$ is 16–256. Fixed, dense, and independent of how many nodes you have.

::: {.fragment}
And the representation usually ends up **meaning something** — which is the part
that still surprises me.
:::
:::
::: {.column width="54%"}
```{r}
#| fig-width: 6
#| fig-height: 4.8
Z <- spectral_embedding(A, d = 2)
par(mar = c(4, 4, 1, 1))
plot(Z[, 1], Z[, 2], type = "n", xlab = "dimension 1", ylab = "dimension 2")
abline(v = 0, col = "grey85", lty = 2)
text(Z[, 1], Z[, 2], labels = seq_len(nrow(Z)), cex = 0.9, col = fac_col)
```

Colour = the faction they actually joined. **The algorithm never saw the
labels.** Axis 1 alone gets 33 of 34 right.
:::
:::

## Everyone has seen this one {.smaller}

$$
\overrightarrow{\text{king}} \;-\; \overrightarrow{\text{man}} \;+\; \overrightarrow{\text{woman}} \;\approx\; \overrightarrow{\text{queen}}
$$

::: {.fragment}
Nobody asked for this. Nobody added a "royalty" column. It fell out of a model
trained to predict nearby words — which means the space has **directions that
correspond to concepts**.
:::

::: {.fragment}
::: {.callout-warning appearance="minimal"}
Obligatory honesty: this example is somewhat cherry-picked. The original code
excludes the input words from the answer, and if you don't exclude them, the
nearest vector to *king − man + woman* is very often just **king**.

The phenomenon is real. The demo is polished. We'll reproduce it on a graph in
about eight slides, with the same caveat attached.
:::
:::

## Why not just one-hot? {.smaller}

**Genuine question for the room: how else would you feed a node to a model?**

::: {.fragment}
One-hot encoding. Node 7 becomes $(0,0,0,0,0,0,1,0,\ldots)$.
:::

::: {.fragment}
It works. It is also the least informative thing you could possibly write down:

- $n$ columns wide, so it changes shape when someone joins
- $\|x_i - x_j\| = \sqrt{2}$ for **every** pair $i \ne j$
- so every node is exactly as far from every other node — **no geometry at all**
:::

::: {.fragment .highlight}
An embedding carries far more than a 1 and a pile of 0s.
:::

## This is really a question about basis {.smaller}

::: {.columns}
::: {.column width="50%"}
A **basis** is a minimal set of vectors that spans the space: every point is
exactly one weighted sum of them.

$$\mathbf{x} = c_1\mathbf{b}_1 + c_2\mathbf{b}_2 + \cdots + c_d\mathbf{b}_d$$

Changing basis does not change the space. **It changes the coordinates** — and
it changes which questions are easy.
:::
::: {.column width="50%"}
::: {.fragment}
One-hot **is** a basis: the standard one, $\mathbf{e}_1 \ldots \mathbf{e}_n$.
One axis per node. Maximally uninformative, and $n$-dimensional.

Spectral embedding is also a basis: eigenvectors of the Laplacian, keeping the
$d$ smoothest.
:::
:::
:::

::: {.fragment .highlight}
An embedding is a **choice of basis**. We would like a good one.
:::

::: {.fragment}
"Good" = few coordinates, and the ones you keep line up with things you care
about.
:::

## Word to vector {.smaller}

**word2vec** (Mikolov et al., 2013): predict a word from the words around it.
Throw away the prediction. Keep the weights. The weights are the embedding.

::: {.fragment}
The Stanford one is **GloVe** (Pennington, Socher & Manning, 2014). Different
route: build the full word co-occurrence matrix, then factorise it.
:::

::: {.fragment}
::: {.callout-note appearance="minimal"}
Hold on to that pairing — *sample the neighbourhood* versus *factorise the
co-occurrence matrix*. It comes back in six slides, and it turns out they are
the same algorithm.
:::
:::

::: {.fragment}
Both rest on one sentence, Firth 1957:

> You shall know a word by the company it keeps.
:::

## Now do it for a graph {.smaller}

::: {.columns}
::: {.column width="52%"}
We need something that maps graph structure — nodes and edges — to vectors.
Unsurprisingly there are many ways.

::: {.fragment}
And there **is** a direct analogue of word2vec. The trick is almost insultingly
simple:
:::

::: {.fragment .highlight}
A random walk on a graph is a sentence.
:::

::: {.fragment}
Take random walks. Write down the nodes you visit, in order. You now have a text
corpus whose "words" are nodes.

Run word2vec on it, unmodified. That is **DeepWalk** (2014).
:::
:::
::: {.column width="48%"}
```r
walks <- random_walks(A, num_walks = 10,
                      walk_length  = 40)
pairs <- walks_to_pairs(walks, window = 5)
fit   <- skipgram_sgns(pairs, n, dim = 32)
```

```{r}
#| fig-width: 5
#| fig-height: 4.2
Z_dw <- deepwalk(A, dim = 32, epochs = 5, seed = 1)
px <- prcomp(Z_dw)$x[, 1:2]
par(mar = c(4, 4, 2, 1))
plot(px, type = "n", xlab = "PC1", ylab = "PC2",
     main = sprintf("5-NN accuracy: %.2f", knn_accuracy(Z_dw, lab, 5)))
text(px[, 1], px[, 2], labels = seq_len(nrow(A)), cex = 0.85, col = fac_col)
```

**~60 lines of base R.** Nothing is $O(n^2)$, which is why this family scaled to
billions of edges.
:::
:::

## The other route: factorise something {.smaller}

Want connected nodes close together? Penalise every edge by the gap it stretches:

$$
\tfrac{1}{2}\sum_{i,j} A_{ij}(f_i - f_j)^2 \;=\; \mathbf{f}^\top \underbrace{(D - A)}_{\text{Laplacian } L} \mathbf{f}
$$

Minimise that, constrain it so it can't collapse to a constant, and you get **the
eigenvectors of $L$ with the smallest eigenvalues** — the smoothest non-constant
functions on the graph.

```r
L  <- diag(rowSums(A)) - A
ev <- eigen(L, symmetric = TRUE)
Z  <- ev$vectors[, (n-1):(n-d)]     # drop the trivial constant vector
```

::: {.fragment}
**Free bonus:** a gap in the eigenvalues tells you how many communities there are.

```{r}
g3 <- sbm_graph(n_per = 40, k = 3, p_in = 0.14, p_out = 0.01, seed = 42)
round(sort(eigen(graph_laplacian(as_adj_dense(g3), "sym"), symmetric = TRUE)$values)[1:6], 3)
```

Three near zero, then a jump. Three communities. Nobody told it three.
:::

## They're all the same algorithm {.smaller}

::: {.callout-important appearance="minimal"}
**DeepWalk is not really a neural method.** Skip-gram provably converges to a
factorisation of a PMI matrix (Levy & Goldberg 2014), and for random walks that
matrix is closed-form (Qiu et al. 2018):

$$
M = \frac{\mathrm{vol}(G)}{bT}\Big(\sum_{r=1}^{T} P^r\Big) D^{-1}, \qquad P = D^{-1}A
$$
:::

::: {.columns}
::: {.column width="50%"}
```{r}
tab <- data.frame(
  Method = c("Spectral", "DeepWalk / NetMF", "node2vec", "LINE"),
  `Factorises` = c("$I - D^{-1/2}AD^{-1/2}$",
                   "$\\log\\max(M, 1)$",
                   "$M$ with 2nd-order $P$",
                   "$M$ with $T = 1$"),
  check.names = FALSE
)
knitr::kable(tab)
```

The same *sample-or-factorise* pairing as word2vec and GloVe — and the same
punchline: it doesn't matter which you pick.
:::
::: {.column width="50%"}
```{r}
t_dw <- system.time(Zd <- deepwalk(A, dim = 32, epochs = 5, seed = 1))[["elapsed"]]
t_nm <- system.time(Zn <- netmf(A, dim = 32))[["elapsed"]]
knitr::kable(data.frame(
  ` ` = c("DeepWalk (sampled)", "NetMF (SVD)"),
  seconds = signif(c(t_dw, t_nm), 3),
  `5-NN acc` = c(knn_accuracy(Zd, lab, 5), knn_accuracy(Zn, lab, 5)),
  check.names = FALSE
))
```

::: {.fragment}
Same answer. **~1000× faster.**

Before you reach for a library: **try an SVD.**
:::
:::
:::

## When $A$ isn't enough: GNNs {.smaller}

Everything so far is a function of the adjacency matrix **alone**. So it cannot
use node features, and it cannot embed a node that wasn't there at training time.

::: {.fragment}
GNNs replace the lookup table with a **function of the neighbourhood**:

$$
H^{(l+1)} = \sigma\big(\hat{A}\,H^{(l)}\,W^{(l)}\big), \qquad \hat{A} = \tilde D^{-1/2}(A+I)\tilde D^{-1/2}
$$

The weights are **shared**, so embedding a brand-new node is just a forward pass.
:::

::: {.fragment}
Two things worth knowing:

- A GCN layer is a **learned smoother**. Stack twenty and every node converges to
  the same vector — which is why GNNs are 2–3 layers deep and everyone is oddly
  quiet about that.
- **A graph attention layer is self-attention masked by the adjacency matrix.**
  That's the whole idea.
:::

# Part 2 {.center}

### Let's actually do it in R

## A drug knowledge graph {.smaller}

```{r}
#| label: kg
#| include: false
gk <- drug_kg()
Ak <- as_adj_dense(gk)
set.seed(1)
Zk <- deepwalk(Ak, dim = 32, num_walks = 40, walk_length = 20, epochs = 20, seed = 1)
rownames(Zk) <- V(gk)$name
```

::: {.columns}
::: {.column width="48%"}
24 drugs, each joined to its **molecular target**, its **pharmacologic class**,
and the **conditions it treats** — which are in turn joined to a body system.

```{r}
knitr::kable(as.data.frame(table(Type = V(gk)$type)))
```

`r vcount(gk)` nodes, `r ecount(gk)` edges, and **no drug is ever joined
directly to another drug.** Every similarity on the next two slides has to
travel through the structure.
:::
::: {.column width="52%"}
```{r}
#| fig-width: 6
#| fig-height: 5
tcol <- c(drug = blue, target = green, class = orange,
          indication = "#8172B2", system = "#937860")
set.seed(11)
par(mar = c(0, 0, 0, 0))
plot(gk, layout = layout_with_fr(gk),
     vertex.color = tcol[V(gk)$type], vertex.frame.color = NA,
     vertex.size = ifelse(V(gk)$type == "drug", 6, 9),
     vertex.label = ifelse(V(gk)$type == "drug", NA, V(gk)$name),
     vertex.label.cex = 0.6, vertex.label.color = "black",
     vertex.label.dist = 0.8, edge.color = "grey80")
```
:::
:::

## Does it know what a drug class is? {.smaller}

Nearest neighbours of `metoprolol`, by cosine similarity:

```{r}
knitr::kable(t(round(nearest(Zk, Zk["metoprolol", ], 5, "metoprolol"), 3)))
```

::: {.fragment}
Its two fellow beta blockers, then the class node, then the receptor all three
of them bind.
:::

::: {.fragment .highlight}
It recovered "beta blocker" without ever being told drug classes exist.
:::

::: {.fragment}
```r
gk <- drug_kg()
Zk <- deepwalk(as_adj_dense(gk), dim = 32, num_walks = 40,
               walk_length = 20, epochs = 20, seed = 1)
rownames(Zk) <- V(gk)$name
nearest(Zk, Zk["metoprolol", ], k = 5, exclude = "metoprolol")
```
:::

## king − man + woman, for drugs {.smaller}

I was genuinely curious whether this would work. Same arithmetic — the question
is whether *"this drug, but for a different disease"* is a **direction**.

::: {.columns}
::: {.column width="50%"}
```r
analogy(Zk, "atorvastatin",
            "hyperlipidemia",
            "depression")
```

> *atorvastatin is to high cholesterol as \_\_\_\_ is to depression*

```{r}
knitr::kable(t(round(analogy(Zk, "atorvastatin", "hyperlipidemia", "depression", 5), 3)))
```

::: {.fragment}
**Fluoxetine.** Prozac. Top of the list — then sertraline and citalopram, with
the serotonin transporter all three of them bind wedged in between.
:::
:::
::: {.column width="50%"}
::: {.fragment}
```r
analogy(Zk, "lisinopril",
            "hypertension",
            "type_2_diabetes")
```

```{r}
knitr::kable(t(round(analogy(Zk, "lisinopril", "hypertension", "type_2_diabetes", 5), 3)))
```

**Metformin.** First-line therapy for exactly that.
:::
:::
:::

::: {.fragment .highlight}
Nobody put a "first-line therapy" column in this graph.
:::

## How to read that {.smaller}

::: {.columns}
::: {.column width="42%"}
**No pharmacology required.** Four facts, and they are all you need:

- **hyperlipidemia** = high cholesterol; **atorvastatin** is the standard drug for it (you may know it as Lipitor)
- **fluoxetine** is the standard drug for **depression** (Prozac)
- the two drugs are chemically unrelated and act on different organs
- the graph contains **only** three kinds of edge: drug→target, drug→class, drug→condition

::: {.fragment}
The blue arrow is one single idea: **"is the standard drug for."**

Start at a disease, follow the arrow, land on its drug.
:::

::: {.fragment}
Now **pick that same arrow up and move it** so its tail sits on a different disease --- depression --- and look at where the head lands.
:::

::: {.fragment .highlight}
It lands on Prozac.
:::

::: {.fragment}
That phrase --- *"is the standard drug for"* --- is written **nowhere** in the graph. It is a direction the geometry worked out on its own, and it points the same way in cardiology and in psychiatry.
:::
:::
::: {.column width="58%"}
```{r}
#| fig-width: 7.2
#| fig-height: 5.9
analogy_plot(Zk, "atorvastatin", "hyperlipidemia", "depression", quiet = TRUE)
```

::: {.fragment}
Two honest notes on the picture:

- The parallelogram is **exact**, not a projection artefact --- the plane is built from the arithmetic itself, so all four corners are drawn where they really are.
- The ranking is by **cosine**, an angle measured at the origin, which is not in this plane. So the winner is not the dot that *looks* nearest the ✕. Hence the labelled link.
:::
:::
:::

## Before anyone gets excited {.smaller}

Two analogies is an anecdote. So here is **every** (drug, its indication, other
indication) triple in the graph — 193 of them — scored on whether the top hit is
a drug that really does treat the third term:

::: {.fragment}
| | |
|---|---|
| top-1 correct | **45%** |
| picking a drug at random | ~12% |

Well above chance. Nowhere near the two I put on the previous slide.
:::

::: {.incremental}
- The graph is **clean by construction**. I wrote it. Every class has exactly
  three interchangeable drugs — precisely the parallel structure the
  parallelogram needs. Real vocabularies are not this tidy.
- 55 nodes, so `analogy()` chooses among ~50 candidates and node type rules out
  most of them before cosine similarity does any work.
- It excludes its own input terms, exactly as the original word2vec evaluation
  code does. **That exclusion is load-bearing.** Comment it out and the answer is
  frequently just the drug you started with.
:::

::: {.fragment}
::: {.callout-note appearance="minimal"}
What I would actually claim: the geometry is doing something real — it recovered
drug class from pure structure — and analogies are a fast, cheap way to
**inspect** an embedding.

Not to make clinical decisions with.
:::
:::

::: {.fragment .highlight}
Run it yourself: `examples/`, base R, no dependency past igraph.
:::

# Part 3 {.center}

### The Claude Code part

## One thing to understand {.smaller}

::: {.fragment .highlight}
You are managing a fixed context window, all the time.
:::

::: {.fragment}
That's the whole mental model. Everything else follows from it.

It is not a database and it is not a colleague with a memory. It is a very good
reasoner looking at **a fixed-size window of text that you chose**.
:::

::: {.fragment}
So the entire skill is: **be deliberate about what's in the window.** Which means
loading things *dynamically* — not everything, all the time.
:::

## `CLAUDE.md` vs. skills {.smaller}

::: {.columns}
::: {.column width="50%"}
### `CLAUDE.md`
> *"Add this to the beginning of every conversation."*

Always loaded. Costs you window on every single turn.

Good for: how you like code written, house style, the two commands that are
never in the README.

::: {.fragment}
Bad for: everything else. A 400-line `CLAUDE.md` is a tax you pay per message
and mostly do not use.
:::
:::
::: {.column width="50%"}
### Skills
> *"Load this when it's relevant."*

A folder of instructions plus a one-line description. The description is always
visible; **the body is only pulled in when it's needed.**

Good for: the Snowflake auth dance, your study-package conventions, the
deployment runbook.

::: {.fragment}
This is the actual unlock. Not *more* context — **the right context, on demand.**
:::
:::
:::

::: {.fragment}
Memory works at both levels too: global, across every project, and local to one.
:::

## Why the terminal changes it {.smaller}

A chatbot can tell you what to type.

::: {.fragment .highlight}
This thing reads files, writes files, and runs commands.
:::

::: {.fragment}
Which sounds incremental and isn't, because it closes the loop: it can **check
its own work.** Write the code, run the code, read the error, fix the code — with
no human in the middle copy-pasting tracebacks.
:::

::: {.fragment}
Every number on the last fifteen slides was produced that way. Including the ones
that came out worse than I hoped, which are still on the slides.
:::

## Things I have actually done with it {.smaller}

::: {.columns}
::: {.column width="50%"}
- **Reproduced a paper.** Multi-armed bandits, from the PDF to a working
  simulation that matched the published figures.
- **Written a course.** This one — a Quarto book, eight chapters, every number
  computed at render time.
- **Lecture scripts and image prompts**, then the ElevenLabs API, then an audio
  course I listen to on the way in.
:::
::: {.column width="50%"}
- **Taught myself things while walking around**, which still feels vaguely
  illegal.
- The general shape: it **adds context to anything**, and it organises and
  connects ideas that were sitting in different places in my head.
:::
:::

::: {.fragment}
::: {.callout-tip appearance="minimal"}
### The pattern, if you take one thing
**Open-ended conversation → structured markdown file → a project around it.**

The conversation is where you work out what you think. The markdown file is where
it stops evaporating. The project is where it has to survive contact with a
compiler.
:::
:::

# Part 4 {.center}

### An idea Donald and I had

## A language with no visible syntax {.smaller}

Donald told me about a programming language. He called it something I probably
shouldn't say into a microphone.

::: {.fragment}
It's actually called **Whitespace** — Edwin Brady and Chris Morris, 2003. The
entire syntax is **space, tab, linefeed**. Every visible character is ignored,
which means a program can hide inside perfectly ordinary text.
:::

::: {.fragment}
It is stack-based and Turing complete: arithmetic, heap storage, subroutines,
jumps, character and numeric I/O.

And when you open a file you see **nothing**. It isn't obfuscated and it isn't
minified — it is illegible in the strict sense. There is nothing there to read.
:::

::: {.fragment}
::: {.callout-note appearance="minimal"}
Pedantic correction I have already received, and deserved: it is **not** "all
spaces". Tabs and linefeeds are load-bearing. A language of nothing but literal
spaces would need some external distinction — run lengths, say — to encode
anything at all.

Three characters. Which is two more than I was hoping for.
:::
:::

::: {.fragment .highlight}
So naturally we started thinking about what to write in it.
:::

## Choosing the target {.smaller}

We wanted the smallest program that is **absolutely critical** to every
tech-forward company on earth and that **nobody has opened in six years**.

::: {.incremental}
- Not the payments code. People read the payments code.
- Not auth. People *definitely* read auth.
- Something that returns the same three bytes, forever, to a caller that is not
  a human.
:::

::: {.fragment .highlight}
The health check endpoint.
:::

::: {.fragment}
It returns `OK`. Every load balancer, orchestrator, uptime monitor and deploy
gate on the planet is downstream of it. It was written by an intern. Its last
commit message is "fix".
:::

::: {.fragment}
**If it stops returning `OK`, the whole company is down** — and the file has 45
bytes in it.
:::

## So we wrote it {.smaller}

```{r}
#| label: ws
#| include: false
ws_path <- if (file.exists("examples/healthcheck.ws")) "examples/healthcheck.ws" else "../examples/healthcheck.ws"
ws_raw <- readChar(ws_path, file.size(ws_path), useBytes = TRUE)
```

::: {.columns}
::: {.column width="46%"}
`examples/healthcheck.ws`, exactly as shown by every editor, reviewer and diff
tool you own:

```{r}
#| comment: ""
cat(ws_raw)
```

::: {.fragment}
That is the entire program. It is not a placeholder.
:::
:::
::: {.column width="54%"}
::: {.fragment}
The same `r nchar(ws_raw, type = "bytes")` bytes, with `·` for space and `→` for
tab:

```{r}
#| comment: ""
cat(gsub("\n", "$\n", gsub("\t", "→", gsub(" ", "·", ws_raw))))
```

Push 79, print it. Push 75, print it. Push 10, print it. Exit.
:::

::: {.fragment}
```
$ python3 scripts/whitespace.py
output when run: 'OK\n'
health check: PASS
```
:::
:::
:::

::: {.fragment}
And because every visible character is ignored, you don't even need a suspicious
blank file. You can paste those 45 bytes **into the trailing whitespace of the
README** and it still runs.
:::

::: {.fragment .highlight}
Ship it on a Friday. Nobody finds it for years.
:::

## Takeaways {.smaller}

::: {.incremental}
1. **Graph embeddings turn structure into geometry.** A graph is already a metric
   space; what it lacks is *coordinates*. An embedding is a choice of basis, and
   a good one puts the things you care about on the axes.

2. **They are all the same algorithm** — a low-rank factorisation of a matrix
   built from $A$. Sample it or factorise it; same answer, and the SVD is a
   thousand times faster. Try an SVD before a library, and always run *"count the
   mutual friends"* as a baseline. It wins more often than anyone admits.

3. **The geometry means something.** Nearest neighbours found drug classes and
   analogies found first-line therapies, out of nothing but edges. Stay
   suspicious of how clean my example was.

4. **Manage the context window.** `CLAUDE.md` is always-on, skills are on-demand,
   and the terminal is what lets it check its own work.
:::

## {.center}

::: {.callout-note appearance="minimal"}
Book, code, slides, the drug graph, and yes, the health check:

### **github.com/ablack3/graph-embeddings-talk**
:::

::: {.fragment .highlight}
Thank you. Please don't stand in the bike lane.
:::
